---
title: "Aging Brain MI Link Visualization"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

This report visualizes inferred meta-interaction (MI) links between individual cells in the MERFISH ageing mouse brain dataset. It generates the circular MI overview used in panel a of the SpiderNet ageing-brain main figure.

The two input tables are exported by `AgingBrain_training_analysis.ipynb`. The edge table is already restricted upstream to directed neighboring cell pairs for which at least one MI strength is at least 0.7; `interaction_num` records how many MI dimensions meet that threshold. This report does not repeat that scientific filtering step.

The workflow includes:

1. define dataset paths and plotting options;
2. load cell metadata and the upstream-filtered directed MI edges;
3. optionally subset the data;
4. compute each cell's incident MI count from the inferred graph;
5. prepare circular layout coordinates;
6. draw cell-type sectors, degree bars, and MI links;
7. export the figure to PDF.

# 1. User settings

Edit this section first when applying the notebook to a new dataset.

```{r user-settings}
# Main input/output directory. Override the environment variable when results are
# stored elsewhere; the default preserves the directory used for the paper analysis.
file_savepath_main <- Sys.getenv(
  "SPIDERNET_AGINGBRAIN_RESULT_DIR",
  unset = "D:/SpiderNet/Results/AgingBrain/V1/SpiderNet_Result_dim30/"
)

# Input files (expected under file_savepath_main by default)
# cell_meta_all.csv: celltype, age, and cellname for each cell
# MI_edge_index_all_choose.csv: source, target, and interaction_num for each
# upstream-selected directed neighboring cell pair
cell_meta_file <- file.path(file_savepath_main, "cell_meta_all.csv")
edge_file      <- file.path(file_savepath_main, "MI_edge_index_all_choose.csv")

# Output figure
output_pdf <- file.path(file_savepath_main, "MI_links.pdf")

# Display options
if_showdifftype    <- FALSE   # keep only cross-celltype edges
if_showTcell       <- TRUE    # keep only edges involving T cell
focus_celltype     <- "T cell"
if_sample          <- FALSE   # optionally downsample cells for quick preview
sample_fraction    <- 0.1
if_transform_degree <- TRUE   # cap degree for visualization
degree_cap          <- 10

# Celltype order on the circle
celltypes <- c(
  "Neuron-Excitatory", "Neuron-Inhibitory", "Neuron-MSN", "Astrocyte",
  "Microglia", "Oligodendrocyte", "OPC", "Endothelial", "Pericyte",
  "VSMC", "VLMC", "Ependymal", "Neuroblast", "NSC", "Macrophage",
  "T cell", "Neutrophil", "B cell"
)

# Celltype colors
color_set <- c(
  "#258B42", "#9FCF89", "#98CB3B", "#4F6AB5", "#6BCCD7", "#8ACEEC",
  "#3FB7E7", "#EB2024", "#7F2321", "#F67F70", "#D25C5A", "#909190",
  "#F7A45C", "#F48B20", "#7D287C", "#F321F8", "#9750B4", "#C176AE"
)
```

# 2. Load packages

```{r packages}
library(circlize)
library(dplyr)
library(tidyr)
library(readr)
```

# 3. Helper functions

These small functions make the workflow easier to reuse on another dataset.

```{r helper-functions}
shorten_ct <- function(ct) {
  mapping <- c(
    "Neuron-Excitatory" = "Neu-Exc.",
    "Neuron-Inhibitory" = "Neu-Inh.",
    "Neuron-MSN"        = "Neu-MSN.",
    "Oligodendrocyte"   = "Oligo.",
    "Neutrophil"        = "Neut.",
    "Macrophage"        = "Macro.",
    "Ependymal"         = "Epend.",
    "Endothelial"       = "Endo.",
    "Neuroblast"        = "Nblast.",
    "Astrocyte"         = "Astro.",
    "Microglia"         = "Micro.",
    "Pericyte"          = "Peri."
  )
  ifelse(ct %in% names(mapping), mapping[ct], ct)
}

read_mi_data <- function(cell_meta_file, edge_file) {
  cell_meta <- read_csv(cell_meta_file, show_col_types = FALSE) %>% as.data.frame()
  edge_df   <- read_csv(edge_file, show_col_types = FALSE) %>% as.data.frame()

  stopifnot("cellname" %in% colnames(cell_meta))
  stopifnot(all(c("source", "target") %in% colnames(edge_df)))

  rownames(cell_meta) <- cell_meta$cellname
  list(cell_meta = cell_meta, edge_df = edge_df)
}

filter_edges_and_cells <- function(cell_meta, edge_df,
                                   if_showdifftype = FALSE,
                                   if_showTcell = FALSE,
                                   focus_celltype = "T cell",
                                   if_sample = FALSE,
                                   sample_fraction = 0.1) {
  cell_meta_ori <- cell_meta

  # Restrict to cells that are present in the edge list
  keep_cells <- unique(c(edge_df$source, edge_df$target))
  cell_meta  <- cell_meta[keep_cells, , drop = FALSE]

  if (if_showTcell) {
    edge_df <- edge_df[
      (cell_meta_ori[edge_df$source, "celltype"] == focus_celltype) |
      (cell_meta_ori[edge_df$target, "celltype"] == focus_celltype),
      , drop = FALSE
    ]
    keep_cells <- unique(c(edge_df$source, edge_df$target))
    cell_meta  <- cell_meta[keep_cells, , drop = FALSE]
  }

  if (if_showdifftype) {
    edge_df <- edge_df[
      cell_meta_ori[edge_df$source, "celltype"] != cell_meta_ori[edge_df$target, "celltype"],
      , drop = FALSE
    ]
    keep_cells <- unique(c(edge_df$source, edge_df$target))
    cell_meta  <- cell_meta[keep_cells, , drop = FALSE]
  }

  if (if_sample && nrow(cell_meta) > 0) {
    n_sample <- max(1, floor(nrow(cell_meta) * sample_fraction))
    sampled_idx <- sample(seq_len(nrow(cell_meta)), n_sample)
    sampled_cells <- cell_meta$cellname[sampled_idx]

    cell_meta <- cell_meta[sampled_idx, , drop = FALSE]
    cell_meta_ori <- cell_meta_ori[sampled_cells, , drop = FALSE]
    edge_df <- edge_df[
      edge_df$source %in% sampled_cells & edge_df$target %in% sampled_cells,
      , drop = FALSE
    ]
  }

  list(cell_meta = cell_meta, cell_meta_ori = cell_meta_ori, edge_df = edge_df)
}

compute_degree_table <- function(cell_meta, cell_meta_ori, edge_df,
                                 if_transform_degree = TRUE,
                                 degree_cap = 10) {
  if (!"interaction_num" %in% colnames(edge_df)) {
    stop("edge_df must contain an 'interaction_num' column.")
  }

  degree_count <- edge_df %>%
    pivot_longer(
      cols = c("source", "target"),
      names_to = "role",
      values_to = "cellname"
    ) %>%
    group_by(cellname) %>%
    summarise(degree = sum(interaction_num, na.rm = TRUE), .groups = "drop")

  degree_count_full <- data.frame(
    cellname = cell_meta_ori$cellname,
    degree   = 0
  )
  degree_count_full$degree <- degree_count$degree[
    match(degree_count_full$cellname, degree_count$cellname)
  ]
  degree_count_full$degree[is.na(degree_count_full$degree)] <- 0

  if (if_transform_degree) {
    degree_count_full$degree <- ifelse(
      degree_count_full$degree > degree_cap,
      degree_cap,
      degree_count_full$degree
    )
  }

  cell_meta <- cell_meta %>%
    left_join(degree_count_full, by = "cellname")
  cell_meta$degree[is.na(cell_meta$degree)] <- 0

  cell_meta
}

prepare_cell_layout <- function(cell_meta, celltypes, min_width = 10) {
  cell_meta <- cell_meta %>%
    group_by(celltype) %>%
    mutate(cell_index = row_number()) %>%
    ungroup()

  cell_counts <- table(cell_meta$celltype)
  scaled_counts <- sqrt(cell_counts)
  scaled_counts <- pmax(scaled_counts, min_width)

  # Ensure all requested celltypes exist in the vector used by circos.initialize
  scaled_counts_full <- setNames(rep(min_width, length(celltypes)), celltypes)
  scaled_counts_full[names(scaled_counts)] <- scaled_counts

  list(
    cell_meta = cell_meta,
    cell_counts = cell_counts,
    scaled_counts = scaled_counts_full
  )
}

draw_mi_circos <- function(cell_meta, edge_df, celltypes, celltype_colors,
                           output_pdf, focus_celltype = "T cell",
                           degree_cap = 10) {
  cairo_pdf(
    filename = output_pdf,
    width = 30,
    height = 30,
    family = "Arial"
  )

  par(family = "Arial")

  circos.clear()
  circos.par(
    start.degree = 180,
    gap.degree = 2,
    cell.padding = c(0, 0, 0, 0),
    track.margin = c(0.01, 0.01),
    canvas.xlim = c(-1.05, 1.05),
    canvas.ylim = c(-1.05, 1.05)
  )

  celltypes_present <- intersect(celltypes, unique(cell_meta$celltype))
  if (length(celltypes_present) == 0) {
    stop("No requested celltypes are present in cell_meta.")
  }

  scaled_counts <- attr(cell_meta, "scaled_counts")
  circos.initialize(
    factors = celltypes_present,
    xlim = cbind(
      rep(0, length(celltypes_present)),
      scaled_counts[celltypes_present]
    )
  )

  # Outer ring
  circos.trackPlotRegion(
    ylim = c(0, 1),
    track.height = 0.05,
    bg.border = NA,
    panel.fun = function(x, y) {}
  )

  for (ct in celltypes_present) {
    highlight.sector(
      sector.index = ct,
      track.index = 1,
      col = adjustcolor(celltype_colors[ct], alpha.f = 1.0),
      border = celltype_colors[ct],
      lwd = 1,
      text = NULL
    )

    xlim <- get.cell.meta.data("xlim", sector.index = ct)
    xcenter <- mean(xlim)
    circos.text(
      x = xcenter,
      y = 2.5,
      labels = shorten_ct(gsub("\\.", " ", ct)),
      sector.index = ct,
      facing = "bending.outside",
      niceFacing = TRUE,
      cex = 7,
      col = "black"
    )
  }

  # Degree bar track
  max_degree <- max(cell_meta$degree)
  cell_meta$scaled_index <- NA

  circos.trackPlotRegion(
    ylim = c(0, max_degree),
    track.height = 0.15,
    bg.border = NA,
    panel.fun = function(x, y) {
      sector <- get.cell.meta.data("sector.index")
      cells  <- cell_meta[cell_meta$celltype == sector, , drop = FALSE]

      if (nrow(cells) > 0) {
        n_real   <- nrow(cells)
        n_scaled <- get.cell.meta.data("xlim", sector.index = sector)[2]
        x_pos <- (cells$cell_index / n_real) * n_scaled

        cell_meta[cell_meta$cellname %in% cells$cellname, "scaled_index"] <<- x_pos

        circos.rect(
          xleft   = x_pos - (0.5 * n_scaled / n_real),
          ybottom = 0,
          xright  = x_pos + (0.5 * n_scaled / n_real),
          ytop    = cells$degree,
          sector.index = sector,
          col     = adjustcolor(celltype_colors[sector], alpha.f = 0.8),
          border  = NA
        )
      }
    }
  )

  # Y axis
  axis_breaks <- unique(c(0, 2, 4, 6, 8, degree_cap))
  axis_labels <- as.character(axis_breaks)
  axis_labels[length(axis_labels)] <- paste0(">", degree_cap)
  axis_sector <- if (focus_celltype %in% celltypes_present) focus_celltype else celltypes_present[1]

  circos.yaxis(
    side = "left",
    at = axis_breaks,
    labels = axis_labels,
    labels.cex = 1.5,
    tick.length = 0.5,
    sector.index = axis_sector,
    track.index = get.current.track.index()
  )

  for (ct in celltypes_present) {
    circos.lines(
      x = get.cell.meta.data("xlim", sector.index = ct),
      y = c(0, 0),
      sector.index = ct,
      col = "black",
      lwd = 0.8
    )
  }

  # Links
  cell2info <- cell_meta %>%
    select(cellname, celltype, scaled_index)

  edge_index_plot <- edge_df %>%
    left_join(cell2info, by = c("source" = "cellname")) %>%
    rename(source_type = celltype, source_index = scaled_index) %>%
    left_join(cell2info, by = c("target" = "cellname")) %>%
    rename(target_type = celltype, target_index = scaled_index)

  apply(edge_index_plot, 1, function(e) {
    s_type <- e[["source_type"]]
    t_type <- e[["target_type"]]
    s_idx  <- as.numeric(e[["source_index"]])
    t_idx  <- as.numeric(e[["target_index"]])

    if (is.na(s_idx) || is.na(t_idx)) return(NULL)
    if (!(s_type %in% celltypes_present) || !(t_type %in% celltypes_present)) return(NULL)

    col_s <- celltype_colors[s_type]
    h_val <- ifelse(s_type == t_type, 0.2, 0.8)

    circos.link(
      sector.index1 = s_type, point1 = s_idx,
      sector.index2 = t_type, point2 = t_idx,
      col    = adjustcolor(col_s, alpha.f = 0.3),
      border = NA,
      h      = h_val,
      lwd    = 0.1
    )
  })

  dev.off()

  invisible(cell_meta)
}
```

# 4. Read input data

```{r read-data}
start_time <- Sys.time()

mi_data <- read_mi_data(
  cell_meta_file = cell_meta_file,
  edge_file = edge_file
)

cell_meta <- mi_data$cell_meta
edge_index_all_choose <- mi_data$edge_df

cat("Loaded cells:", nrow(cell_meta), "\n")
cat("Loaded edges:", nrow(edge_index_all_choose), "\n")
```

# 5. Apply optional filtering

This step keeps only the cells and edges that will be displayed.

```{r filter-data}
filtered <- filter_edges_and_cells(
  cell_meta = cell_meta,
  edge_df = edge_index_all_choose,
  if_showdifftype = if_showdifftype,
  if_showTcell = if_showTcell,
  focus_celltype = focus_celltype,
  if_sample = if_sample,
  sample_fraction = sample_fraction
)

cell_meta <- filtered$cell_meta
cell_meta_ori <- filtered$cell_meta_ori
edge_index_all_choose <- filtered$edge_df

cat("Cells after filtering:", nrow(cell_meta), "\n")
cat("Edges after filtering:", nrow(edge_index_all_choose), "\n")
```

# 6. Compute incident MI count

For each cell, the displayed degree is the sum of `interaction_num` across all incident edges. Thus, it summarizes the number of above-threshold MI dimensions on those edges rather than the sum of the underlying continuous MI strengths.

```{r compute-degree}
cell_meta <- compute_degree_table(
  cell_meta = cell_meta,
  cell_meta_ori = cell_meta_ori,
  edge_df = edge_index_all_choose,
  if_transform_degree = if_transform_degree,
  degree_cap = degree_cap
)

summary(cell_meta$degree)
```

# 7. Prepare circular layout

Here we assign cells to within-celltype positions and compute scaled sector widths.

```{r prepare-layout}
layout_info <- prepare_cell_layout(
  cell_meta = cell_meta,
  celltypes = celltypes,
  min_width = 10
)

cell_meta <- layout_info$cell_meta
scaled_counts <- layout_info$scaled_counts
cell_counts <- layout_info$cell_counts

celltype_colors <- setNames(color_set[seq_along(celltypes)], celltypes)

print(cell_counts)

# Store scaled_counts as an attribute so the plotting function can access it
attr(cell_meta, "scaled_counts") <- scaled_counts
```

# 8. Draw and save the circular plot

```{r make-plot}
draw_mi_circos(
  cell_meta = cell_meta,
  edge_df = edge_index_all_choose,
  celltypes = celltypes,
  celltype_colors = celltype_colors,
  output_pdf = output_pdf,
  focus_celltype = focus_celltype,
  degree_cap = degree_cap
)

cat("Saved figure to:", output_pdf, "\n")
```

# 9. Session summary

```{r summary}
end_time <- Sys.time()
cat("Finished.\n")
cat("Elapsed time:", round(difftime(end_time, start_time, units = "secs"), 2), "seconds\n")
```

# Notes for adapting this notebook

When applying this report to a new dataset, the sections that usually need to change are:

- `SPIDERNET_AGINGBRAIN_RESULT_DIR` (or `file_savepath_main`), input file names, and output file name;
- the target celltype used for focused plotting;
- the `celltypes` order;
- the color palette;
- any filtering condition specific to the dataset.

The core functions for reading, filtering, degree calculation, layout preparation, and plotting can usually remain unchanged as long as the input tables use the same column names.
