---
title: "GO Enrichment Visualization for MI-29"
author: "Junjie Tang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# Overview

This report visualizes the selected Gene Ontology (GO) biological process enrichment results for MI-29-associated sender regulators and receiver targets in the MERFISH ageing mouse brain dataset. It exports the two individual dot columns and a connected sender-receiver overview.

The enrichment calculations and term selection are performed upstream in `AgingBrain_training_analysis.ipynb`. The input CSV files therefore contain precomputed enrichment statistics for the terms selected for the published panel; this report does not rerun GO enrichment or multiple-testing correction.

It generates:

- **p1**: GO enrichment of top regulators (sender side)
- **p2**: GO enrichment of top targets (receiver side)
- **connected_plot**: both columns with the manuscript's two schematic program-group ribbons

The workflow is organized into four steps:

1. Load required packages and input files.
2. Clean the displayed term labels and derive plotting quantities from the stored statistics.
3. Define the final term order for the sender and receiver plots.
4. Generate the individual and connected plots as PNG/PDF files, with plotting data and ribbon-group CSVs.

Ribbons represent the manuscript's program groupings. Their width is a layout choice, not an inferred interaction strength or a statistical test. Bubble sizes retain the shared scale derived from the original adjusted P values.

# Load packages and define paths

```{r packages-and-paths}
library(ggplot2)
library(dplyr)
library(stringr)

result_dir <- Sys.getenv(
  "SPIDERNET_AGINGBRAIN_RESULT_DIR",
  unset = "D:/SpiderNet/Results/AgingBrain/V1/SpiderNet_Result_dim30"
)

toptarget_file <- file.path(result_dir, "GO_enrichment_toptarget_MI29.csv")
topregulator_file <- file.path(result_dir, "GO_enrichment_topregulator_MI29.csv")
```

# Read input tables

Each table is expected to contain `Term` and `Adjusted P-value`; base R converts the latter to `Adjusted.P.value` when reading the CSV file.

```{r read-data}
GO_enrichment_toptarget_MI29 <- read.csv(toptarget_file)
GO_enrichment_topregulator_MI29 <- read.csv(topregulator_file)
```

# Define helper function for data cleaning

```{r helper-function}
prepare_go_df <- function(df) {
  df %>%
    mutate(
      # Remove GO accession IDs such as "(GO:XXXXXXX)" from the term labels.
      Term_clean = str_replace(Term, "\\s*\\(GO:\\d+\\)", ""),
      # Convert adjusted p-values to -log10 scale for plotting.
      logP = -log10(Adjusted.P.value)
    )
}
```

# Prepare sender and receiver enrichment tables

```{r prepare-data}
df_sender <- prepare_go_df(GO_enrichment_topregulator_MI29)
df_receiver <- prepare_go_df(GO_enrichment_toptarget_MI29)

# Define the final display order for the sender enrichment plot (p1).
order_p1 <- rev(c(
  "regulation of interferon-gamma production",
  "interferon-gamma-mediated signaling pathway",
  "positive regulation of T cell cytokine production",
  "cytokine-mediated signaling pathway",
  "T cell activation",
  "cellular senescence",
  "positive regulation of cellular senescence"
))

# Define the final display order for the receiver enrichment plot (p2).
order_p2 <- rev(c(
  "inflammatory response",
  "antigen receptor-mediated signaling pathway",
  "positive regulation of leukocyte cell-cell adhesion",
  "nucleotide-excision repair, DNA incision",
  "global genome nucleotide-excision repair"
))

df_sender <- df_sender %>%
  filter(Term_clean %in% order_p1) %>%
  mutate(Term_clean = factor(Term_clean, levels = order_p1))

df_receiver <- df_receiver %>%
  filter(Term_clean %in% order_p2) %>%
  mutate(Term_clean = factor(Term_clean, levels = order_p2))
```

# Define shared plotting scale

```{r shared-scale}
# Use a shared point-size scale so that p1 and p2 are visually comparable.
all_logP <- c(df_sender$logP, df_receiver$logP)
min_logP <- min(all_logP)
max_logP <- max(all_logP)

size_breaks <- c(2, 5, 8)
```

# Plot p1: sender-side GO enrichment

```{r plot-p1, fig.width=15, fig.height=7}
p1 <- ggplot(df_sender, aes(x = 1, y = Term_clean)) +
  geom_point(aes(size = logP), color = "#FD2D2D") +
  scale_size(
    range = c(7, 15),
    limits = c(min_logP, max_logP),
    breaks = size_breaks,
    name = "-log10(P)"
  ) +
  scale_y_discrete(position = "left") +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 27, lineheight = 0.9)
  )

p1
```

```{r save-p1, eval=TRUE}
ggsave(
  file.path(result_dir, "GO_enrichment_topregulator_MI29_SingleColumn_BubblePlot.png"),
  plot = p1,
  width = 15,
  height = 7,
  dpi = 300
)

ggsave(
  file.path(result_dir, "GO_enrichment_topregulator_MI29_SingleColumn_BubblePlot.pdf"),
  plot = p1,
  width = 15,
  height = 7,
  dpi = 300
)
```

# Plot p2: receiver-side GO enrichment

```{r plot-p2, fig.width=12, fig.height=7}
p2 <- ggplot(df_receiver, aes(x = 1, y = Term_clean)) +
  geom_point(aes(size = logP), color = "#FD2D2D") +
  scale_size(
    range = c(7, 15),
    limits = c(min_logP, max_logP),
    breaks = size_breaks,
    name = "-log10(P)"
  ) +
  scale_y_discrete(position = "right") +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 27, lineheight = 0.9)
  )

p2
```

```{r save-p2, eval=TRUE}
ggsave(
  file.path(result_dir, "GO_enrichment_toptarget_MI29_SingleColumn_BubblePlot.png"),
  plot = p2,
  width = 12,
  height = 7,
  dpi = 300
)

ggsave(
  file.path(result_dir, "GO_enrichment_toptarget_MI29_SingleColumn_BubblePlot.pdf"),
  plot = p2,
  width = 12,
  height = 7,
  dpi = 300
)
```

# Connected sender-receiver programs

The first five sender terms and first three receiver terms form the immune/cytokine
group; the remaining two terms on each side form the senescence/DNA-repair group.
Term selection, order, adjusted P values, bubble colour and size scale are shared
with the individual plots. Only label wrapping and the combined layout differ.

```{r connected-programs}
build_program_column <- function(df, term_order, side, x, source_csv) {
  ordered_terms <- rev(term_order)
  if (anyDuplicated(as.character(df$Term_clean)) ||
      !setequal(as.character(df$Term_clean), ordered_terms)) {
    stop(paste(side, "GO terms are missing or duplicated; cannot draw a complete connected panel."))
  }
  column <- df[match(ordered_terms, as.character(df$Term_clean)), , drop = FALSE]
  if (any(!is.finite(column$logP))) {
    stop(paste(side, "contains non-finite enrichment sizes."))
  }
  column$Side <- side
  column$DisplayOrder <- seq_len(nrow(column))
  column$GO_ID <- str_extract(column$Term, "GO:[0-9]+")
  column$DisplayTerm <- str_wrap(as.character(column$Term_clean), width = 34)
  column$x <- x
  column$y <- seq(6, 0, length.out = nrow(column))
  column$ProgramGroup <- ifelse(column$DisplayOrder <= if (side == "Sender") 5 else 3,
                                "immune", "senescence_repair")
  column$SourceCSV <- basename(source_csv)
  column$SourceMD5 <- unname(tools::md5sum(source_csv))
  column
}

connected_data <- bind_rows(
  build_program_column(df_sender, order_p1, "Sender", 0, topregulator_file),
  build_program_column(df_receiver, order_p2, "Receiver", 1.8, toptarget_file)
)

group_connections <- data.frame(
  ProgramGroup = c("immune", "senescence_repair"),
  Description = c("Immune/cytokine processes and neighboring-cell immune responses",
                  "Cellular senescence and neighboring-cell DNA repair"),
  Meaning = "Manuscript-defined schematic grouping; ribbon width is not a statistic",
  stringsAsFactors = FALSE
)
for (side in c("Sender", "Receiver")) {
  group_connections[[paste0(side, "_GO_IDs")]] <- vapply(
    group_connections$ProgramGroup,
    function(group) paste(connected_data$GO_ID[
      connected_data$Side == side & connected_data$ProgramGroup == group], collapse = ";"),
    character(1)
  )
}

program_ribbon <- function(group, sender_top, sender_bottom, receiver_top, receiver_bottom, bend) {
  t <- seq(0, 1, length.out = 120)
  smooth <- t^2 * (3 - 2 * t)
  x <- 0.35 + (1.45 - 0.35) * t
  top <- sender_top + (receiver_top - sender_top) * smooth - bend * 4 * t * (1 - t)
  bottom <- sender_bottom + (receiver_bottom - sender_bottom) * smooth + bend * 4 * t * (1 - t)
  data.frame(x = c(x, rev(x)), y = c(top, rev(bottom)), ProgramGroup = group)
}
ribbons <- bind_rows(
  program_ribbon("immune", 6.40, 1.60, 6.40, 2.35, 0.50),
  program_ribbon("senescence_repair", 1.40, -0.40, 2.15, -0.40, 0.22)
)

connected_plot <- ggplot() +
  geom_polygon(data = ribbons, aes(x, y, group = ProgramGroup, fill = ProgramGroup),
               show.legend = FALSE) +
  scale_fill_manual(values = c(immune = "#f7c0c1", senescence_repair = "#b8acd0")) +
  annotate("rect", xmin = c(-0.35, 1.45), xmax = c(0.35, 2.15),
           ymin = -0.45, ymax = 6.45, fill = NA, colour = "#555555", linewidth = 0.4) +
  geom_point(data = connected_data, aes(x, y, size = logP), colour = "#FD2D2D") +
  geom_text(data = filter(connected_data, Side == "Sender"),
            aes(x = x - 0.51, y = y, label = DisplayTerm),
            hjust = 1, size = 3.8, lineheight = 0.95) +
  geom_text(data = filter(connected_data, Side == "Receiver"),
            aes(x = x + 0.51, y = y, label = DisplayTerm),
            hjust = 0, size = 3.8, lineheight = 0.95) +
  annotate("text", x = c(0, 1.8), y = 6.94,
           label = c("MI-29 sender", "MI-29 receiver"), size = 4.0) +
  scale_size(range = c(7, 15), limits = c(min_logP, max_logP), breaks = size_breaks,
             name = "Enrichment significance\n-log10(adjusted P)") +
  coord_cartesian(xlim = c(-4.9, 6.7), ylim = c(-0.65, 7.20), expand = FALSE, clip = "off") +
  labs(title = "MI-29 molecular organization") +
  theme_void(base_family = "sans") +
  theme(plot.title = element_text(hjust = 0.5, size = 15, margin = margin(b = 12)),
        legend.position = "bottom", legend.direction = "horizontal",
        legend.title = element_text(size = 10), legend.text = element_text(size = 10),
        plot.margin = margin(12, 12, 12, 12))

connected_stem <- file.path(result_dir, "GO_enrichment_MI29_SenderReceiver_Connected_BubblePlot")
ggsave(paste0(connected_stem, ".pdf"), connected_plot, width = 12, height = 7.2)
ggsave(paste0(connected_stem, ".png"), connected_plot, width = 12, height = 7.2,
       dpi = 300, bg = "white")
write.csv(connected_data, paste0(connected_stem, "_data.csv"), row.names = FALSE)
write.csv(group_connections, paste0(connected_stem, "_connections.csv"), row.names = FALSE)
```

# Session information

```{r session-info}
sessionInfo()
```
