---
title: "GO Enrichment Visualization for MI29"
author: "Junjie Tang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# Overview

This R Markdown document generates two final GO enrichment bubble plots for MI29:

- **p1**: GO enrichment of top regulators (sender side)
- **p2**: GO enrichment of top targets (receiver side)

The workflow is organized into four steps:

1. Load required packages and input files.
2. Clean and standardize the GO enrichment tables.
3. Define the final term order for the sender and receiver plots.
4. Generate and save the two final figures.

Only the final figures `p1` and `p2` are kept in this version. Intermediate figures from the original script have been removed.

# Load packages and define paths

```{r packages-and-paths}
library(ggplot2)
library(dplyr)
library(stringr)

result_dir <- "D:/SpiderNet/Results/AgingBrain/V1/SpiderNet_Result_dim30"

toptarget_file <- file.path(result_dir, "GO_enrichment_toptarget_MI29.csv")
topregulator_file <- file.path(result_dir, "GO_enrichment_topregulator_MI29.csv")
```

# Read input tables

```{r read-data}
GO_enrichment_toptarget_MI29 <- read.csv(toptarget_file)
GO_enrichment_topregulator_MI29 <- read.csv(topregulator_file)
```

# Define helper function for data cleaning

```{r helper-function}
prepare_go_df <- function(df) {
  df %>%
    mutate(
      # Remove GO accession IDs such as "(GO:XXXXXXX)" from the term labels.
      Term_clean = str_replace(Term, "\\s*\\(GO:\\d+\\)", ""),
      # Extract the number of overlapping genes from strings such as "5/120".
      GeneCount = as.numeric(sub("/.*", "", Overlap)),
      # Convert adjusted p-values to -log10 scale for plotting.
      logP = -log10(Adjusted.P.value)
    )
}
```

# Prepare sender and receiver enrichment tables

```{r prepare-data}
df_sender <- prepare_go_df(GO_enrichment_topregulator_MI29)
df_receiver <- prepare_go_df(GO_enrichment_toptarget_MI29)

# Define the final display order for the sender enrichment plot (p1).
order_p1 <- rev(c(
  "regulation of interferon-gamma production",
  "interferon-gamma-mediated signaling pathway",
  "positive regulation of T cell cytokine production",
  "cytokine-mediated signaling pathway",
  "T cell activation",
  "cellular senescence",
  "positive regulation of cellular senescence"
))

# Define the final display order for the receiver enrichment plot (p2).
order_p2 <- rev(c(
  "inflammatory response",
  "antigen receptor-mediated signaling pathway",
  "positive regulation of leukocyte cell-cell adhesion",
  "nucleotide-excision repair, DNA incision",
  "global genome nucleotide-excision repair"
))

df_sender <- df_sender %>%
  filter(Term_clean %in% order_p1) %>%
  mutate(Term_clean = factor(Term_clean, levels = order_p1))

df_receiver <- df_receiver %>%
  filter(Term_clean %in% order_p2) %>%
  mutate(Term_clean = factor(Term_clean, levels = order_p2))
```

# Define shared plotting scale

```{r shared-scale}
# Use a shared point-size scale so that p1 and p2 are visually comparable.
all_logP <- c(df_sender$logP, df_receiver$logP)
min_logP <- min(all_logP)
max_logP <- max(all_logP)

size_breaks <- c(2, 5, 8)
```

# Plot p1: sender-side GO enrichment

```{r plot-p1, fig.width=7.8, fig.height=7}
p1 <- ggplot(df_sender, aes(x = 1, y = Term_clean)) +
  geom_point(aes(size = logP), color = "#FD2D2D") +
  scale_size(
    range = c(7, 15),
    limits = c(min_logP, max_logP),
    breaks = size_breaks,
    name = "-log10(P)"
  ) +
  scale_y_discrete(position = "left") +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 27, lineheight = 0.9)
  )

p1
```

```{r save-p1, eval=TRUE}
ggsave(
  file.path(result_dir, "GO_enrichment_topregulator_MI29_SingleColumn_BubblePlot.png"),
  plot = p1,
  width = 7.8,
  height = 7,
  dpi = 300
)

ggsave(
  file.path(result_dir, "GO_enrichment_topregulator_MI29_SingleColumn_BubblePlot.pdf"),
  plot = p1,
  width = 7.8,
  height = 7,
  dpi = 300
)
```

# Plot p2: receiver-side GO enrichment

```{r plot-p2, fig.width=12, fig.height=7}
p2 <- ggplot(df_receiver, aes(x = 1, y = Term_clean)) +
  geom_point(aes(size = logP), color = "#FD2D2D") +
  scale_size(
    range = c(7, 15),
    limits = c(min_logP, max_logP),
    breaks = size_breaks,
    name = "-log10(P)"
  ) +
  scale_y_discrete(position = "right") +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 27, lineheight = 0.9)
  )

p2
```

```{r save-p2, eval=TRUE}
ggsave(
  file.path(result_dir, "GO_enrichment_toptarget_MI29_SingleColumn_BubblePlot.png"),
  plot = p2,
  width = 12,
  height = 7,
  dpi = 300
)

ggsave(
  file.path(result_dir, "GO_enrichment_toptarget_MI29_SingleColumn_BubblePlot.pdf"),
  plot = p2,
  width = 12,
  height = 7,
  dpi = 300
)
```

# Session information

```{r session-info}
sessionInfo()
```
