---
title: "SpiderNet MI Circos Visualization"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This report generates circos visualizations for SpiderNet meta-interaction outputs in the AgingBrain dataset. The current workflow focuses on three related plots for **MI-29**:

1. A full cell-type interaction plot with node labels.
2. A full cell-type interaction plot without labels.
3. A T-cell-centered interaction plot colored by effect direction and significance.

The report assumes that the input CSV files have already been exported to the SpiderNet result directory.

# Setup

```{r libraries}
library(circlize)
library(dplyr)
library(scales)
```

```{r paths}
result_dir <- "D:/SpiderNet/Results/AgingBrain/V1/SpiderNet_Result_dim30/"

all_links_file <- file.path(result_dir, "MI-29_all_links.csv")
tcell_links_file <- file.path(result_dir, "MI-29_Tcell_links.csv")

output_all_labeled <- file.path(result_dir, "MI-29_all_CIRCOS_links_nodesColored_arrow.pdf")
output_all_no_label <- file.path(result_dir, "MI-29_all_CIRCOS_links_nodesColored_arrow_no_label.png")
output_tcell_signif <- file.path(result_dir, "MI29_Tcell_CIRCOS_links_nodesColored_arrow_significantColored.pdf")
```

# Shared definitions

```{r shared-definitions}
celltypes <- c(
  "Neuron-Excitatory", "Neuron-Inhibitory", "Neuron-MSN", "Astrocyte", "Microglia", "Oligodendrocyte",
  "OPC", "Endothelial", "Pericyte", "VSMC", "VLMC", "Ependymal", "Neuroblast", "NSC", "Macrophage",
  "Neutrophil", "T cell", "B cell"
)

celltype_colors <- c(
  "Neuron-Excitatory" = "#258B42",
  "Neuron-Inhibitory" = "#9FCF89",
  "Neuron-MSN" = "#98CB3B",
  "Astrocyte" = "#4F6AB5",
  "Microglia" = "#6BCCD7",
  "Oligodendrocyte" = "#8ACEEC",
  "OPC" = "#3FB7E7",
  "Endothelial" = "#EB2024",
  "Pericyte" = "#7F2321",
  "VSMC" = "#F67F70",
  "VLMC" = "#D25C5A",
  "Ependymal" = "#909190",
  "Neuroblast" = "#F7A45C",
  "NSC" = "#F48B20",
  "Macrophage" = "#7D287C",
  "Neutrophil" = "#9750B4",
  "T cell" = "#F321F8",
  "B cell" = "#C176AE"
)

shorten_ct <- function(ct) {
  dplyr::case_when(
    ct == "Neuron-Excitatory" ~ "Neu-Exc.",
    ct == "Neuron-Inhibitory" ~ "Neu-Inh.",
    ct == "Neuron-MSN" ~ "Neu-MSN.",
    ct == "Oligodendrocyte" ~ "Oligo.",
    ct == "Neutrophil" ~ "Neut.",
    ct == "Macrophage" ~ "Macro.",
    ct == "Ependymal" ~ "Epend.",
    ct == "Endothelial" ~ "Endo.",
    ct == "Neuroblast" ~ "Nblast.",
    ct == "Astrocyte" ~ "Astro.",
    ct == "Microglia" ~ "Micro.",
    ct == "Pericyte" ~ "Peri.",
    TRUE ~ ct
  )
}

initialize_circos_layout <- function(nodes, with_padding = TRUE) {
  circos.clear()
  circos.par(
    gap.degree = 4,
    start.degree = 90,
    track.margin = c(0.01, 0.01),
    cell.padding = if (with_padding) c(0.02, 0, 0.02, 0) else c(0, 0, 0, 0)
  )
  circos.initialize(
    factors = nodes,
    xlim = cbind(rep(0, length(nodes)), rep(1, length(nodes)))
  )
}

add_sector_labels <- function(nodes, cex = 1.4) {
  circos.trackPlotRegion(
    ylim = c(0, 1),
    track.height = 0.08,
    bg.border = NA,
    panel.fun = function(x, y) {
      sector <- get.cell.meta.data("sector.index")
      circos.text(
        x = 0.5,
        y = 2,
        labels = shorten_ct(sector),
        facing = "bending.outside",
        niceFacing = TRUE,
        cex = cex
      )
    }
  )
}

add_invisible_track <- function() {
  circos.trackPlotRegion(
    ylim = c(0, 1),
    track.height = 0.001,
    bg.border = NA,
    panel.fun = function(x, y) {}
  )
}

draw_node_points <- function(nodes, point_size = 3.5) {
  for (nd in nodes) {
    circos.points(
      x = 0.5,
      y = 0,
      sector.index = nd,
      pch = 21,
      cex = point_size,
      bg = celltype_colors[nd],
      col = "black"
    )
  }
}

draw_directed_links <- function(df) {
  for (i in seq_len(nrow(df))) {
    lw <- df$link_width[i]
    circos.link(
      df$Sender[i], 0.5,
      df$Receiver[i], 0.5,
      col = df$link_col[i],
      border = NA,
      lwd = lw,
      directional = 1,
      arr.length = 0.3,
      arr.width = 0.3,
      arr.lwd = lw * 0.4
    )
  }
}
```

# Full MI-29 network across all cell-type pairs

This section visualizes the full MI-29 interaction network. Link darkness and width both reflect interaction strength, and stronger links are plotted last so they remain visible on top.

```{r all-links-data}
df_all <- read.csv(all_links_file)

df_all$Mean[is.na(df_all$Mean)] <- 0
df_all$Mean <- df_all$Mean^1.3

df_all <- df_all %>%
  arrange(Mean)

df_all$link_width <- scales::rescale(df_all$Mean, to = c(0.01, 3))
width_floor <- quantile(df_all$link_width, 0.1)
width_cutoff <- quantile(df_all$link_width, 0.8)
df_all$link_width <- ifelse(df_all$link_width < width_cutoff, width_floor, df_all$link_width)
df_all$link_col <- grey(1 - scales::rescale(df_all$Mean, to = c(0, 1)))

nodes_all <- intersect(celltypes, unique(c(df_all$Sender, df_all$Receiver)))
```

## Labeled version

```{r all-links-labeled, eval=FALSE}
pdf(output_all_labeled, width = 10, height = 10)

initialize_circos_layout(nodes_all)
add_sector_labels(nodes_all, cex = 1.4)
draw_directed_links(df_all)
draw_node_points(nodes_all, point_size = 3.5)

circos.clear()
dev.off()
```

## No-label version

This version is useful when the figure is intended for composition in a multi-panel layout and text labels are added elsewhere.

```{r all-links-no-label, eval=FALSE}
png(output_all_no_label, width = 1800, height = 1800, res = 300)

initialize_circos_layout(nodes_all, with_padding = FALSE)
add_invisible_track()
draw_directed_links(df_all)
draw_node_points(nodes_all, point_size = 2.2)

circos.clear()
dev.off()
```

# T-cell-centered MI-29 network

This section focuses on T-cell outgoing links. Significant edges are colored by correlation, while non-significant edges are shown in gray.

```{r tcell-data}
df_tcell <- read.csv(tcell_links_file)

df_tcell$Corr[is.na(df_tcell$Corr)] <- 0
df_tcell$P.value[is.na(df_tcell$P.value)] <- 1
df_tcell$Sender <- "T cell"

df_tcell$link_width <- scales::rescale(df_tcell$Mean, to = c(1, 6))

col_fun <- colorRamp2(
  c(-0.4, 0, 0.4),
  c("#5463FF", "#ECECEC", "#FF1818")
)

df_tcell$link_col <- ifelse(
  df_tcell$P.value >= 0.05,
  "#BFBFBF",
  col_fun(df_tcell$Corr)
)

nodes_tcell <- intersect(celltypes, unique(c(df_tcell$Sender, df_tcell$Receiver)))
```

```{r tcell-plot, eval=FALSE}
cairo_pdf(
  filename = output_tcell_signif,
  width = 10,
  height = 10,
  family = "Arial"
)
par(family = "Arial")

initialize_circos_layout(nodes_tcell)
add_sector_labels(nodes_tcell, cex = 1.4)

for (i in seq_len(nrow(df_tcell))) {
  lw <- df_tcell$link_width[i]
  this_col <- if (df_tcell$P.value[i] >= 0.05) {
    adjustcolor("#BFBFBF", alpha.f = 0.8)
  } else {
    adjustcolor(df_tcell$link_col[i], alpha.f = 1.0)
  }

  circos.link(
    df_tcell$Sender[i], 0.5,
    df_tcell$Receiver[i], 0.5,
    col = this_col,
    border = NA,
    lwd = lw,
    directional = 1,
    arr.length = 0.3,
    arr.width = 0.3,
    arr.lwd = lw * 0.4
  )
}

draw_node_points(nodes_tcell, point_size = 3.5)

circos.clear()
dev.off()
```

# Notes

- Update `result_dir` if the MI export files are stored in a different SpiderNet result folder.
- To visualize another MI, replace the input CSV file names and adjust the output file names accordingly.
- The plotting chunks are set to `eval = FALSE` by default so the document can be reviewed without regenerating files automatically. Set them to `TRUE` when you are ready to export the figures.
