---
title: "PerturbFISH GO Enrichment Visualization"
output: html_notebook
---

This notebook reorganizes the provided GO enrichment plotting script into a cleaner, easier-to-follow workflow. It focuses on loading GO enrichment results for one meta-interaction (MI), preparing sender and receiver enrichment tables, and generating single-column bubble plots for publication-style visualization.

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)

library(ggplot2)
library(dplyr)
library(stringr)
```

## 2. User settings

Edit this section first when switching to another result folder or MI.

```{r user-settings}
###############################################################################
# Main inputs
###############################################################################

# Main result directory
RESULT_DIR <- "D:/SpiderNet/Results/PerturbFISH/V1/SpiderNet_Result_dim23"

# Meta-interaction of interest
# MI_OI <- "MI10"
# # MI_OI <- "MI8"
MI_OI <- "MI17"
# MI_OI <- "MI13"

###############################################################################
# Plot settings
###############################################################################

BASIC_BUBBLE_LOW_COLOR  <- "#8CC7FF"
BASIC_BUBBLE_HIGH_COLOR <- "#004DA8"
SINGLE_COLUMN_COLOR     <- "#FD2D2D"

# Whether to display GO IDs in the single-column plots
SHOW_GO_ID_IN_SINGLE_COLUMN <- TRUE

###############################################################################
# Optional manual term order
# Set to NULL to use automatic ordering by adjusted p-value
###############################################################################

SENDER_TERM_ORDER   <- NULL
RECEIVER_TERM_ORDER <- NULL

# if (MI_OI == "MI10") {
#   SENDER_TERM_ORDER <- rev(c(
#     "response to interleukin-1 (GO:0070555)",
#     "cellular response to lipopolysaccharide (GO:0071222)",
#     "positive regulation of interleukin-8 production (GO:0032757)",
#     "positive regulation of leukocyte chemotaxis (GO:0002690)",
#     "CD40 signaling pathway (GO:0023035)",
#     "regulation of protein phosphorylation (GO:0001932)"
#   ))
# 
#   RECEIVER_TERM_ORDER <- rev(c(
#     "cellular response to cytokine stimulus (GO:0071345)",
#     "regulation of protein phosphorylation (GO:0001932)",
#     "positive regulation of protein kinase activity (GO:0045860)",
#     "integrin-mediated signaling pathway (GO:0007229)",
#     "leukocyte cell-cell adhesion (GO:0007159)",
#     "lamellipodium assembly (GO:0030032)",
#     "CD40 signaling pathway (GO:0023035)",
#     "positive regulation of protein kinase B signaling (GO:0051897)"
#   ))
# } else if (MI_OI == "MI8") {
#   SENDER_TERM_ORDER <- rev(c(
#     "I-kappaB kinase/NF-kappaB signaling (GO:0007249)",
#     "TRIF-dependent toll-like receptor signaling pathway (GO:0035666)",
#     "transmembrane receptor protein tyrosine kinase signaling pathway (GO:0007169)",
#     "phosphatidylinositol 3-kinase signaling (GO:0014065)"
#   ))
# 
#   RECEIVER_TERM_ORDER <- rev(c(
#     "type I interferon signaling pathway (GO:0060337)",
#     "antigen processing and presentation of peptide antigen via MHC class I (GO:0002474)",
#     "T cell receptor signaling pathway (GO:0050852)",
#     "positive regulation of protein kinase B signaling (GO:0051897)"
#   ))
# }
if (MI_OI == "MI17") {
  SENDER_TERM_ORDER <- rev(c(
    "cellular response to cytokine stimulus (GO:0071345)",
    "response to interleukin-1 (GO:0070555)",
    "cellular response to reactive oxygen species (GO:0034614)",
    "transforming growth factor beta receptor signaling pathway (GO:0007179)",
    "epithelial to mesenchymal transition (GO:0001837)",
    "extracellular matrix organization (GO:0030198)"
  ))

  RECEIVER_TERM_ORDER <- rev(c(
    "negative regulation of apoptotic process (GO:0043066)",
    "response to endoplasmic reticulum stress (GO:0034976)",
    "cellular response to reactive oxygen species (GO:0034614)",
    "integrin-mediated signaling pathway (GO:0007229)",
    "heterotypic cell-cell adhesion (GO:0034113)",
    "regulation of pathway-restricted SMAD protein phosphorylation (GO:0060393)"
  ))
} else if (MI_OI == "MI13") {
  SENDER_TERM_ORDER <- rev(c(
    "I-kappaB kinase/NF-kappaB signaling (GO:0007249)",
    "TRIF-dependent toll-like receptor signaling pathway (GO:0035666)",
    "transmembrane receptor protein tyrosine kinase signaling pathway (GO:0007169)",
    "phosphatidylinositol 3-kinase signaling (GO:0014065)"
  ))

  RECEIVER_TERM_ORDER <- rev(c(
    "type I interferon signaling pathway (GO:0060337)",
    "antigen processing and presentation of peptide antigen via MHC class I (GO:0002474)",
    "T cell receptor signaling pathway (GO:0050852)",
    "positive regulation of protein kinase B signaling (GO:0051897)"
  ))
}
```

## 3. Helper functions

```{r helper-functions}
build_go_file_paths <- function(result_dir, mi_oi) {
  list(
    target_csv    = file.path(result_dir, paste0("GO_", mi_oi, "_target_enrichment.csv")),
    regulator_csv = file.path(result_dir, paste0("GO_", mi_oi, "_regulator_enrichment.csv"))
  )
}

prepare_go_dataframe <- function(df) {
  df %>%
    mutate(
      Term_clean = str_replace(Term, "\\s*\\(GO:\\d+\\)", ""),
      GeneCount  = as.numeric(sub("/.*", "", Overlap)),
      logP       = -log10(pmax(Adjusted.P.value, .Machine$double.xmin))
    ) %>%
    arrange(Adjusted.P.value)
}

apply_term_order <- function(df,
                             term_order = NULL,
                             use_full_term = FALSE,
                             keep_only_ordered_terms = FALSE) {
  term_col <- if (use_full_term) "Term" else "Term_clean"

  if (is.null(term_order)) {
    df[[term_col]] <- factor(df[[term_col]], levels = unique(df[[term_col]]))
    return(df)
  }

  ordered_terms <- term_order[term_order %in% unique(df[[term_col]])]

  if (keep_only_ordered_terms) {
    df <- df[df[[term_col]] %in% ordered_terms, , drop = FALSE]
    df[[term_col]] <- factor(df[[term_col]], levels = ordered_terms)
  } else {
    remaining_terms <- setdiff(unique(df[[term_col]]), ordered_terms)
    df[[term_col]] <- factor(df[[term_col]], levels = c(ordered_terms, remaining_terms))
  }

  df
}

plot_basic_bubble <- function(df, output_path, low_color, high_color,
                              width = 8, height = 6, dpi = 300) {
  p <- ggplot(df, aes(x = logP, y = Term_clean)) +
    geom_point(aes(size = GeneCount, color = logP)) +
    scale_color_gradient(low = low_color, high = high_color) +
    scale_size(range = c(4, 10)) +
    labs(
      x = "-log10(P value)",
      y = "",
      size = "Gene count",
      color = "-log10(P)"
    ) +
    theme_bw(base_size = 14) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(size = 12),
      axis.title.x = element_text(size = 14)
    )

  print(p)
  ggsave(output_path, plot = p, width = width, height = height, dpi = dpi)
}

plot_single_column_bubble <- function(df,
                                      output_pdf,
                                      point_color,
                                      size_limits,
                                      size_breaks,
                                      y_axis_position = c("left", "right"),
                                      width = 6.5,
                                      height = 7,
                                      dpi = 300) {
  y_axis_position <- match.arg(y_axis_position)

  p <- ggplot(df, aes(x = 1, y = DisplayTerm)) +
    geom_point(aes(size = logP), color = point_color) +
    scale_size(
      range  = c(3, 10),
      limits = size_limits,
      breaks = size_breaks,
      name   = "-log10(P)",
      guide  = guide_legend(
        override.aes = list(
          size = scales::rescale(
            size_breaks,
            to   = c(3, 10),
            from = size_limits
          ),
          color = point_color
        )
      )
    ) +
    scale_y_discrete(position = y_axis_position) +
    labs(x = "", y = "") +
    theme_bw(base_size = 14) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x      = element_blank(),
      axis.ticks.x     = element_blank(),
      axis.text.y      = element_text(size = 27, lineheight = 0.9)
    )

  print(p)
  ggsave(output_pdf, plot = p, width = width, height = height, dpi = dpi)
}
```

## 4. Build file paths and load GO enrichment tables

```{r load-go-tables}
go_paths <- build_go_file_paths(RESULT_DIR, MI_OI)

go_paths
```

```{r read-go-tables}
GO_enrichment_toptarget    <- read.csv(go_paths$target_csv)
GO_enrichment_topregulator <- read.csv(go_paths$regulator_csv)

cat("Receiver table rows:", nrow(GO_enrichment_toptarget), "\n")
cat("Sender table rows:", nrow(GO_enrichment_topregulator), "\n")
```

## 5. Prepare sender and receiver enrichment tables

```{r prepare-data}
df_receiver <- prepare_go_dataframe(GO_enrichment_toptarget)
df_sender   <- prepare_go_dataframe(GO_enrichment_topregulator)
```

```{r preview-data}
head(df_sender)
head(df_receiver)
```

## 6. Prepare data for single-column bubble plots

Use one shared `-log10(adjusted p-value)` scale across sender and receiver so the two plots are directly comparable.

```{r prepare-single-column-data}
all_logP <- c(df_sender$logP, df_receiver$logP)

min_logP <- min(all_logP)
max_logP <- max(all_logP)

size_breaks <- c(
  -log10(0.05),
  -log10(0.01),
  -log10(0.001)
)

size_limits <- range(c(min_logP, max_logP, size_breaks))

if (SHOW_GO_ID_IN_SINGLE_COLUMN) {
  df_sender$DisplayTerm   <- df_sender$Term
  df_receiver$DisplayTerm <- df_receiver$Term
} else {
  df_sender$DisplayTerm   <- df_sender$Term_clean
  df_receiver$DisplayTerm <- df_receiver$Term_clean
}


df_sender <- apply_term_order(
  df = df_sender,
  term_order = SENDER_TERM_ORDER,
  use_full_term = SHOW_GO_ID_IN_SINGLE_COLUMN,
  keep_only_ordered_terms = TRUE
)

df_receiver <- apply_term_order(
  df = df_receiver,
  term_order = RECEIVER_TERM_ORDER,
  use_full_term = SHOW_GO_ID_IN_SINGLE_COLUMN,
  keep_only_ordered_terms = TRUE
)

if (SHOW_GO_ID_IN_SINGLE_COLUMN) {
  df_sender$DisplayTerm   <- factor(df_sender$DisplayTerm, levels = levels(df_sender$Term))
  df_receiver$DisplayTerm <- factor(df_receiver$DisplayTerm, levels = levels(df_receiver$Term))
} else {
  df_sender$DisplayTerm   <- factor(df_sender$DisplayTerm, levels = levels(df_sender$Term_clean))
  df_receiver$DisplayTerm <- factor(df_receiver$DisplayTerm, levels = levels(df_receiver$Term_clean))
}

list(
  sender_terms   = levels(df_sender$DisplayTerm),
  receiver_terms = levels(df_receiver$DisplayTerm),
  size_limits    = size_limits,
  size_breaks    = size_breaks
)
df_sender

```

## 7. Generate single-column bubble plots

These are the main output plots used in the original script.

### 7.1 Sender / regulator plot

```{r plot-sender, fig.width=13, fig.height=7}
# if (MI_OI == "MI10") {
#   width_use = 13
# } else if (MI_OI == "MI8") {
#   width_use = 16
# }
if (MI_OI == "MI17") {
  width_use = 15
} else if (MI_OI == "MI13") {
  width_use = 16
}
plot_single_column_bubble(
  df = df_sender,
  output_pdf = file.path(
    RESULT_DIR,
    paste0("GO_enrichment_topregulator_", MI_OI, "_SingleColumn_BubblePlot.pdf")
  ),
  point_color     = SINGLE_COLUMN_COLOR,
  size_limits     = size_limits,
  size_breaks     = size_breaks,
  y_axis_position = "left",
  width           = width_use,
  height          = 7
)
```

### 7.2 Receiver / target plot

```{r plot-receiver, fig.width=12.8, fig.height=7}
# if (MI_OI == "MI10") {
#   width_use = 12.8
# } else if (MI_OI == "MI8") {
#   width_use = 17
# }
if (MI_OI == "MI17") {
  width_use = 16
} else if (MI_OI == "MI13") {
  width_use = 17
}

plot_single_column_bubble(
  df = df_receiver,
  output_pdf = file.path(
    RESULT_DIR,
    paste0("GO_enrichment_toptarget_", MI_OI, "_SingleColumn_BubblePlot.pdf")
  ),
  point_color     = SINGLE_COLUMN_COLOR,
  size_limits     = size_limits,
  size_breaks     = size_breaks,
  y_axis_position = "right",
  width           = width_use,
  height          = 7
)
```

## 8. Output summary

```{r output-summary}
cat("Main output directory:\n", RESULT_DIR, "\n\n")
cat("Main figures:\n")
cat("- ", file.path(RESULT_DIR, paste0("GO_enrichment_topregulator_", MI_OI, "_SingleColumn_BubblePlot.pdf")), "\n", sep = "")
cat("- ", file.path(RESULT_DIR, paste0("GO_enrichment_toptarget_", MI_OI, "_SingleColumn_BubblePlot.pdf")), "\n", sep = "")
```

## 9. Session info

```{r session-info}
sessionInfo()
```
