---
title: "PerturbFISH GO Enrichment Visualization"
output: html_notebook
---

This notebook visualizes precomputed Gene Ontology (GO) Biological Process enrichment results for a selected PerturbFISH meta-interaction (MI). It reads separate sender-regulator and receiver-target enrichment tables, derives plotting values from gene overlaps and adjusted P values, applies the manuscript term selections and ordering, and writes paired single-column bubble plots. With the default MI-17 configuration, these plots provide the two GO-program columns used in the PerturbFISH regulator-target panel (main-text PerturbFISH figure, panel d); the paired overview, including schematic connecting ribbons, is generated below as well.

The expected input tables are Enrichr-style CSV files named `GO_<MI>_regulator_enrichment.csv` and `GO_<MI>_target_enrichment.csv`. They are generated by the upstream PerturbFISH analysis and must include `Term`, `Overlap`, and `Adjusted P-value` columns. Required R packages are `ggplot2`, `dplyr`, `stringr`, and `scales`.

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)

library(ggplot2)
library(dplyr)
library(stringr)
```

## 2. User settings

Edit this section when switching to another result folder or MI. The results directory can also be supplied through the `SPIDERNET_PERTURBFISH_RESULTS_DIR` environment variable.

```{r user-settings}
###############################################################################
# Main inputs
###############################################################################

# Default result directory used for the manuscript analysis. Set
# SPIDERNET_PERTURBFISH_RESULTS_DIR to run from a different filesystem layout.
DEFAULT_RESULT_DIR <- "output"
RESULT_DIR <- Sys.getenv(
  "SPIDERNET_PERTURBFISH_RESULTS_DIR",
  unset = DEFAULT_RESULT_DIR
)
OUTPUT_DIR <- Sys.getenv("SPIDERNET_PERTURBFISH_PLOT_DIR", unset = "output")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

# Meta-interaction shown in the main-text PerturbFISH figure
# MI_OI <- "MI10"
# # MI_OI <- "MI8"
MI_OI <- "MI17"
# MI_OI <- "MI13"

###############################################################################
# Plot settings
###############################################################################

SINGLE_COLUMN_COLOR     <- "#FD2D2D"

# Whether to display GO IDs in the single-column plots
SHOW_GO_ID_IN_SINGLE_COLUMN <- TRUE

###############################################################################
# Optional manual term order. Set to NULL to retain the order produced after
# sorting the enrichment table by adjusted P value.
###############################################################################

SENDER_TERM_ORDER   <- NULL
RECEIVER_TERM_ORDER <- NULL

# if (MI_OI == "MI10") {
#   SENDER_TERM_ORDER <- rev(c(
#     "response to interleukin-1 (GO:0070555)",
#     "cellular response to lipopolysaccharide (GO:0071222)",
#     "positive regulation of interleukin-8 production (GO:0032757)",
#     "positive regulation of leukocyte chemotaxis (GO:0002690)",
#     "CD40 signaling pathway (GO:0023035)",
#     "regulation of protein phosphorylation (GO:0001932)"
#   ))
# 
#   RECEIVER_TERM_ORDER <- rev(c(
#     "cellular response to cytokine stimulus (GO:0071345)",
#     "regulation of protein phosphorylation (GO:0001932)",
#     "positive regulation of protein kinase activity (GO:0045860)",
#     "integrin-mediated signaling pathway (GO:0007229)",
#     "leukocyte cell-cell adhesion (GO:0007159)",
#     "lamellipodium assembly (GO:0030032)",
#     "CD40 signaling pathway (GO:0023035)",
#     "positive regulation of protein kinase B signaling (GO:0051897)"
#   ))
# } else if (MI_OI == "MI8") {
#   SENDER_TERM_ORDER <- rev(c(
#     "I-kappaB kinase/NF-kappaB signaling (GO:0007249)",
#     "TRIF-dependent toll-like receptor signaling pathway (GO:0035666)",
#     "transmembrane receptor protein tyrosine kinase signaling pathway (GO:0007169)",
#     "phosphatidylinositol 3-kinase signaling (GO:0014065)"
#   ))
# 
#   RECEIVER_TERM_ORDER <- rev(c(
#     "type I interferon signaling pathway (GO:0060337)",
#     "antigen processing and presentation of peptide antigen via MHC class I (GO:0002474)",
#     "T cell receptor signaling pathway (GO:0050852)",
#     "positive regulation of protein kinase B signaling (GO:0051897)"
#   ))
# }
if (MI_OI == "MI17") {
  SENDER_TERM_ORDER <- rev(c(
    "cellular response to cytokine stimulus (GO:0071345)",
    "response to interleukin-1 (GO:0070555)",
    "cellular response to reactive oxygen species (GO:0034614)",
    "transforming growth factor beta receptor signaling pathway (GO:0007179)",
    "epithelial to mesenchymal transition (GO:0001837)",
    "extracellular matrix organization (GO:0030198)"
  ))

  RECEIVER_TERM_ORDER <- rev(c(
    "negative regulation of apoptotic process (GO:0043066)",
    "response to endoplasmic reticulum stress (GO:0034976)",
    "cellular response to reactive oxygen species (GO:0034614)",
    "integrin-mediated signaling pathway (GO:0007229)",
    "heterotypic cell-cell adhesion (GO:0034113)",
    "regulation of pathway-restricted SMAD protein phosphorylation (GO:0060393)"
  ))
} else if (MI_OI == "MI13") {
  SENDER_TERM_ORDER <- rev(c(
    "I-kappaB kinase/NF-kappaB signaling (GO:0007249)",
    "TRIF-dependent toll-like receptor signaling pathway (GO:0035666)",
    "transmembrane receptor protein tyrosine kinase signaling pathway (GO:0007169)",
    "phosphatidylinositol 3-kinase signaling (GO:0014065)"
  ))

  RECEIVER_TERM_ORDER <- rev(c(
    "type I interferon signaling pathway (GO:0060337)",
    "antigen processing and presentation of peptide antigen via MHC class I (GO:0002474)",
    "T cell receptor signaling pathway (GO:0050852)",
    "positive regulation of protein kinase B signaling (GO:0051897)"
  ))
}
```

## 3. Helper functions

`read.csv()` converts the input column name `Adjusted P-value` to `Adjusted.P.value`. The helpers below preserve that convention, parse the gene-overlap numerator, and prepare the plotting order.

```{r helper-functions}
build_go_file_paths <- function(result_dir, mi_oi) {
  list(
    target_csv    = file.path(result_dir, paste0("GO_", mi_oi, "_target_enrichment.csv")),
    regulator_csv = file.path(result_dir, paste0("GO_", mi_oi, "_regulator_enrichment.csv"))
  )
}

prepare_go_dataframe <- function(df) {
  df %>%
    mutate(
      Term_clean = str_replace(Term, "\\s*\\(GO:\\d+\\)", ""),
      GeneCount  = as.numeric(sub("/.*", "", Overlap)),
      logP       = -log10(pmax(Adjusted.P.value, .Machine$double.xmin))
    ) %>%
    arrange(Adjusted.P.value)
}

apply_term_order <- function(df,
                             term_order = NULL,
                             use_full_term = FALSE,
                             keep_only_ordered_terms = FALSE) {
  term_col <- if (use_full_term) "Term" else "Term_clean"

  if (is.null(term_order)) {
    df[[term_col]] <- factor(df[[term_col]], levels = unique(df[[term_col]]))
    return(df)
  }

  ordered_terms <- term_order[term_order %in% unique(df[[term_col]])]

  if (keep_only_ordered_terms) {
    df <- df[df[[term_col]] %in% ordered_terms, , drop = FALSE]
    df[[term_col]] <- factor(df[[term_col]], levels = ordered_terms)
  } else {
    remaining_terms <- setdiff(unique(df[[term_col]]), ordered_terms)
    df[[term_col]] <- factor(df[[term_col]], levels = c(ordered_terms, remaining_terms))
  }

  df
}

plot_single_column_bubble <- function(df,
                                      output_pdf,
                                      point_color,
                                      size_limits,
                                      size_breaks,
                                      y_axis_position = c("left", "right"),
                                      width = 6.5,
                                      height = 7,
                                      dpi = 300) {
  y_axis_position <- match.arg(y_axis_position)

  p <- ggplot(df, aes(x = 1, y = DisplayTerm)) +
    geom_point(aes(size = logP), color = point_color) +
    scale_size(
      range  = c(3, 10),
      limits = size_limits,
      breaks = size_breaks,
      name   = "-log10(P)",
      guide  = guide_legend(
        override.aes = list(
          size = scales::rescale(
            size_breaks,
            to   = c(3, 10),
            from = size_limits
          ),
          color = point_color
        )
      )
    ) +
    scale_y_discrete(position = y_axis_position) +
    labs(x = "", y = "") +
    theme_bw(base_size = 14) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x      = element_blank(),
      axis.ticks.x     = element_blank(),
      axis.text.y      = element_text(size = 27, lineheight = 0.9)
    )

  print(p)
  ggsave(output_pdf, plot = p, width = width, height = height, dpi = dpi)
}
```

## 4. Build file paths and load GO enrichment tables

The input CSV files are produced upstream by `PerturbFISH_insilicospatialperturbation.ipynb`; this notebook does not rerun the enrichment analysis.

```{r load-go-tables}
go_paths <- build_go_file_paths(RESULT_DIR, MI_OI)

go_paths
```

```{r read-go-tables}
GO_enrichment_toptarget    <- read.csv(go_paths$target_csv)
GO_enrichment_topregulator <- read.csv(go_paths$regulator_csv)

cat("Receiver table rows:", nrow(GO_enrichment_toptarget), "\n")
cat("Sender table rows:", nrow(GO_enrichment_topregulator), "\n")
```

## 5. Prepare sender and receiver enrichment tables

For each enriched term, `GeneCount` is the numerator of the Enrichr overlap field and `logP` is the negative base-10 logarithm of the adjusted P value. Values at or below machine precision are bounded only to keep the logarithm finite.

```{r prepare-data}
df_receiver <- prepare_go_dataframe(GO_enrichment_toptarget)
df_sender   <- prepare_go_dataframe(GO_enrichment_topregulator)
```

```{r preview-data}
head(df_sender)
head(df_receiver)
```

## 6. Prepare data for single-column bubble plots

Use one shared `-log10(adjusted P value)` scale across sender and receiver so the two plots are directly comparable. The legacy columns retain their original six-term selections. The paired overview below uses the five terms per side in the supplied manuscript; other significant terms remain in the input tables.

```{r prepare-single-column-data}
all_logP <- c(df_sender$logP, df_receiver$logP)

min_logP <- min(all_logP)
max_logP <- max(all_logP)

size_breaks <- c(
  -log10(0.05),
  -log10(0.01),
  -log10(0.001)
)

size_limits <- range(c(min_logP, max_logP, size_breaks))

if (SHOW_GO_ID_IN_SINGLE_COLUMN) {
  df_sender$DisplayTerm   <- df_sender$Term
  df_receiver$DisplayTerm <- df_receiver$Term
} else {
  df_sender$DisplayTerm   <- df_sender$Term_clean
  df_receiver$DisplayTerm <- df_receiver$Term_clean
}


df_sender <- apply_term_order(
  df = df_sender,
  term_order = SENDER_TERM_ORDER,
  use_full_term = SHOW_GO_ID_IN_SINGLE_COLUMN,
  keep_only_ordered_terms = TRUE
)

df_receiver <- apply_term_order(
  df = df_receiver,
  term_order = RECEIVER_TERM_ORDER,
  use_full_term = SHOW_GO_ID_IN_SINGLE_COLUMN,
  keep_only_ordered_terms = TRUE
)

if (SHOW_GO_ID_IN_SINGLE_COLUMN) {
  df_sender$DisplayTerm   <- factor(df_sender$DisplayTerm, levels = levels(df_sender$Term))
  df_receiver$DisplayTerm <- factor(df_receiver$DisplayTerm, levels = levels(df_receiver$Term))
} else {
  df_sender$DisplayTerm   <- factor(df_sender$DisplayTerm, levels = levels(df_sender$Term_clean))
  df_receiver$DisplayTerm <- factor(df_receiver$DisplayTerm, levels = levels(df_receiver$Term_clean))
}

list(
  sender_terms   = levels(df_sender$DisplayTerm),
  receiver_terms = levels(df_receiver$DisplayTerm),
  size_limits    = size_limits,
  size_breaks    = size_breaks
)
df_sender

```

## 7. Generate single-column bubble plots

These PDFs are the sender-regulator and receiver-target columns used for the PerturbFISH MI-17 program panel. The paired overview below adds curves connecting the manuscript-defined biological process groups.

### 7.1 Sender / regulator plot

```{r plot-sender, fig.width=13, fig.height=7}
# if (MI_OI == "MI10") {
#   width_use = 13
# } else if (MI_OI == "MI8") {
#   width_use = 16
# }
if (MI_OI == "MI17") {
  width_use = 15
} else if (MI_OI == "MI13") {
  width_use = 16
}
plot_single_column_bubble(
  df = df_sender,
  output_pdf = file.path(
    OUTPUT_DIR,
    paste0("GO_enrichment_topregulator_", MI_OI, "_SingleColumn_BubblePlot.pdf")
  ),
  point_color     = SINGLE_COLUMN_COLOR,
  size_limits     = size_limits,
  size_breaks     = size_breaks,
  y_axis_position = "left",
  width           = width_use,
  height          = 7
)
```

### 7.2 Receiver / target plot

```{r plot-receiver, fig.width=12.8, fig.height=7}
# if (MI_OI == "MI10") {
#   width_use = 12.8
# } else if (MI_OI == "MI8") {
#   width_use = 17
# }
if (MI_OI == "MI17") {
  width_use = 16
} else if (MI_OI == "MI13") {
  width_use = 17
}

plot_single_column_bubble(
  df = df_receiver,
  output_pdf = file.path(
    OUTPUT_DIR,
    paste0("GO_enrichment_toptarget_", MI_OI, "_SingleColumn_BubblePlot.pdf")
  ),
  point_color     = SINGLE_COLUMN_COLOR,
  size_limits     = size_limits,
  size_breaks     = size_breaks,
  y_axis_position = "right",
  width           = width_use,
  height          = 7
)
```

## 8. Output summary

```{r output-summary}
cat("Main output directory:\n", RESULT_DIR, "\n\n")
cat("Main figures:\n")
cat("- ", file.path(OUTPUT_DIR, paste0("GO_enrichment_topregulator_", MI_OI, "_SingleColumn_BubblePlot.pdf")), "\n", sep = "")
cat("- ", file.path(OUTPUT_DIR, paste0("GO_enrichment_toptarget_", MI_OI, "_SingleColumn_BubblePlot.pdf")), "\n", sep = "")
```

## 9. Paired sender-receiver program overview

The paired panel uses the five terms per side shown in the supplied manuscript.
The original six-term single-column outputs above remain unchanged. Ribbons show
the two manuscript-defined biological groupings, not estimated edge strengths or
individual statistical tests. Missing manuscript terms stop the paired export
with an explicit error, so stale inputs cannot silently produce an incomplete figure.

```{r paired-programs}
paper_dir <- file.path(OUTPUT_DIR, "paper_figures")
dir.create(paper_dir, recursive = TRUE, showWarnings = FALSE)
program_order <- data.frame(
  Side = rep(c("Sender", "Receiver"), each = 5),
  Order = rep(1:5, 2),
  GO_ID = c("GO:0071345", "GO:0070555", "GO:0034614", "GO:0007179", "GO:0030198",
            "GO:0043066", "GO:0034976", "GO:0034614", "GO:0007229", "GO:0060393"),
  DisplayTerm = c("Cellular response to\ncytokine stimulus", "Response to interleukin-1",
    "Cellular response to\nreactive oxygen species", "TGF-beta receptor\nsignaling pathway",
    "Extracellular\nmatrix organization", "Negative regulation of\napoptotic process",
    "Response to endoplasmic\nreticulum stress", "Cellular response to\nreactive oxygen species",
    "Integrin-mediated\nsignaling pathway", "Regulation of SMAD\nprotein phosphorylation"),
  ProgramGroup = rep(c(rep("Cytokine / oxidative-stress and T-cell stress", 3),
                       rep("ECM / TGF-beta and integrin / SMAD", 2)), 2),
  stringsAsFactors = FALSE
)
paired_input <- bind_rows(
  mutate(prepare_go_dataframe(GO_enrichment_topregulator), Side = "Sender"),
  mutate(prepare_go_dataframe(GO_enrichment_toptarget), Side = "Receiver")
) %>% mutate(GO_ID = str_extract(Term, "GO:[0-9]+"))
if (anyDuplicated(paired_input[c("Side", "GO_ID")])) {
  stop("Duplicate GO IDs within a side; refusing to choose a row silently.")
}
paired_programs <- program_order %>%
  left_join(paired_input, by = c("Side", "GO_ID"))
paired_programs <- paired_programs %>% mutate(
  x = ifelse(Side == "Sender", 0, 1.8), y = 5 - Order,
  Status = ifelse(is.na(Term), "absent_from_saved_enrichment", "available"),
  SourceCSV = ifelse(Side == "Sender", basename(go_paths$regulator_csv), basename(go_paths$target_csv))
)
write.csv(paired_programs, file.path(paper_dir, "paired_GO_programs.csv"), row.names = FALSE)
missing_programs <- paired_programs %>% filter(Status != "available")
write.csv(missing_programs, file.path(paper_dir, "paired_GO_missing_terms.csv"), row.names = FALSE)
if (nrow(missing_programs)) {
  stop(paste(nrow(missing_programs), "manuscript terms absent: check checkpoint provenance and rerun --stage enrichment."))
}

program_ribbon <- function(top, bottom, bend, group) {
  t <- seq(0, 1, length.out = 100)
  x <- 0.19 + (1.61 - 0.19) * t
  data.frame(x = c(x, rev(x)),
             y = c(top - bend * 4 * t * (1 - t), rev(bottom + bend * 4 * t * (1 - t))),
             group = group)
}
ribbons <- bind_rows(program_ribbon(4.35, 1.65, 0.44, "stress"),
                     program_ribbon(1.35, -0.35, 0.24, "matrix"))
write.csv(data.frame(
  Group = unique(program_order$ProgramGroup),
  Sender_GO_IDs = c("GO:0071345;GO:0070555;GO:0034614", "GO:0007179;GO:0030198"),
  Receiver_GO_IDs = c("GO:0043066;GO:0034976;GO:0034614", "GO:0007229;GO:0060393"),
  Meaning = "Manuscript-defined schematic grouping; ribbon width is not a statistic"
), file.path(paper_dir, "paired_GO_connections.csv"), row.names = FALSE)

paired_plot <- ggplot() +
  geom_polygon(data = ribbons, aes(x, y, group = group, fill = group), show.legend = FALSE) +
  scale_fill_manual(values = c(stress = "#f7c0c1", matrix = "#b8acd0")) +
  annotate("rect", xmin = c(-0.18, 1.62), xmax = c(0.18, 1.98),
           ymin = -0.45, ymax = 4.45, fill = NA, colour = "black", linewidth = 0.35) +
  geom_point(data = filter(paired_programs, Status == "available"),
             aes(x, y, size = logP), colour = "#cc1168") +
  geom_text(data = filter(paired_programs, Side == "Sender"),
            aes(x = x - 0.31, y = y, label = DisplayTerm), hjust = 1, size = 3.3, lineheight = 0.95) +
  geom_text(data = filter(paired_programs, Side == "Receiver"),
            aes(x = x + 0.31, y = y, label = DisplayTerm), hjust = 0, size = 3.3, lineheight = 0.95) +
  annotate("text", x = c(0, 1.8), y = 4.85, label = c("Sender program", "Receiver program"), size = 3.7) +
  scale_size(range = c(3, 10), limits = size_limits, breaks = size_breaks, labels = c("1.3", "2", "3"),
             name = "Enrichment significance\n-log10(adjusted P)") +
  coord_cartesian(xlim = c(-3.4, 5.2), ylim = c(-0.65, 5.1), clip = "off") +
  labs(title = "MI-17 molecular organization") +
  theme_void(base_family = "sans") +
  theme(plot.title = element_text(hjust = 0.5, size = 13, margin = margin(b = 10)),
        plot.caption = element_text(hjust = 0.5, size = 9, colour = "#555555"),
        legend.position = "bottom", legend.direction = "horizontal", legend.title = element_text(size = 9),
        plot.margin = margin(12, 12, 10, 12))
ggsave(file.path(paper_dir, "Fig4d.pdf"), paired_plot, width = 10.5, height = 5.3)
ggsave(file.path(paper_dir, "Fig4d.png"), paired_plot, width = 10.5, height = 5.3, dpi = 300, bg = "white")
```

## 10. Session info

```{r session-info}
sessionInfo()
```
