---
title: "Pancancer MI link visualization"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Overview

This notebook visualizes pancancer meta-interaction links in a circular layout.
It is adapted to the pancancer MI-2 link table and keeps the workflow modular so
that the plotting settings, cell-type order, and file paths can be edited in one place.

The workflow includes:

1. define pancancer-specific file paths and plotting options;
2. load the MI-2 sender-receiver summary table;
3. preprocess link weights and keep the desired node order;
4. prepare plotting helpers for node labels, link styling, and self-loops;
5. draw a labeled directional circos plot;
6. draw a no-label directional circos plot for figure assembly.

# 1. User settings

Edit this section first when applying the notebook to a new result directory.

```{r user-settings}
# Main input/output directory
file_savepath_main <- "D:/SpiderNet/Results/Pancancer/V1/SpiderNet_Result_dim11/"

# Input file generated from the Python analysis
mi_link_file <- file.path(file_savepath_main, "MI2_all_links.csv")

# Output figures
output_pdf_labeled  <- file.path(file_savepath_main, "MI2_all_CIRCOS_links_nodesColored_arrow.pdf")
output_png_nolabel  <- file.path(file_savepath_main, "MI2_all_CIRCOS_links_nodesColored_arrow_NO_LABEL.png")

# Link-weight display options
link_power        <- 2
link_width_range  <- c(0.01, 4)
quantile_floor_hi <- 0.90
quantile_floor_lo <- 0.10

# Cell-type order on the circle
celltypes <- c(
  "B cell",
  "Breast-cancercell",
  "CD4_T",
  "CD8_T/NK",
  "Colon-cancercell",
  "DC",
  "Endothelial",
  "Epithelial",
  "Fibroblast",
  "Liver-cancercell",
  "Lung-cancercell",
  "Macrophage",
  "Mast cell",
  "Melanoma-cancercell",
  "Ovarian-cancercell",
  "Prostate-cancercell",
  "Treg",
  "Uterine-cancercell",
  "low_exp"
)

# Cell-type colors
color_set <- c(
  "#8894F1",
  "#7F3480",
  "#FF259E",
  "#FF52DB",
  "#F7B2EF",
  "#2C5E1A",
  "#029991",
  "#E384FF",
  "#944E13",
  "#FF8B6C",
  "#B37166",
  "#74650F",
  "#7228B8",
  "#009D4E",
  "#DD0303",
  "#028DCE",
  "#DE0878",
  "#FAB12F",
  "#BBC58F"
)
```

# 2. Load packages

```{r packages}
library(circlize)
library(dplyr)
library(readr)
library(scales)
```

# 3. Helper functions

These helper functions keep the main plotting section concise and easier to adapt.

```{r helper-functions}
shorten_celltype <- function(ct) {
  mapping <- c(
    "Breast-cancercell"   = "Breast-tumor",
    "Colon-cancercell"    = "Colon-tumor",
    "Liver-cancercell"    = "Liver-tumor",
    "Lung-cancercell"     = "Lung-tumor",
    "Melanoma-cancercell" = "Melanoma",
    "Ovarian-cancercell"  = "Ovarian-tumor",
    "Prostate-cancercell" = "Prostate-tumor",
    "Uterine-cancercell"  = "Uterine-tumor"
  )

  ifelse(ct %in% names(mapping), mapping[ct], ct)
}

read_pancancer_links <- function(mi_link_file) {
  df <- read_csv(mi_link_file, show_col_types = FALSE) %>% as.data.frame()

  stopifnot(all(c("Sender", "Receiver", "Mean") %in% colnames(df)))

  df
}

prepare_link_dataframe <- function(df,
                                   link_power = 3,
                                   link_width_range = c(0.01, 3),
                                   quantile_floor_hi = 0.80,
                                   quantile_floor_lo = 0.10) {
  df_use <- df

  df_use$Mean[is.na(df_use$Mean)] <- 0
  df_use$Mean <- df_use$Mean ^ link_power

  # Draw weak links first and strong links last so stronger links stay on top.
  df_use <- df_use %>% arrange(Mean)

  df_use$link_width <- scales::rescale(df_use$Mean, to = link_width_range)

  q_hi <- quantile(df_use$link_width, quantile_floor_hi, na.rm = TRUE)
  q_lo <- quantile(df_use$link_width, quantile_floor_lo, na.rm = TRUE)
  df_use$link_width <- ifelse(df_use$link_width < q_hi, q_lo, df_use$link_width)

  # Stronger links are shown with darker grayscale.
  df_use$link_col <- grey(1 - scales::rescale(df_use$Mean, to = c(0, 1)))

  df_use
}

get_nodes_in_order <- function(df, celltypes) {
  celltypes[celltypes %in% unique(c(df$Sender, df$Receiver))]
}

initialize_circos_layout <- function(nodes, with_labels = TRUE, label_cex = 1.4) {
  circos.clear()
  circos.par(
    gap.degree = 4,
    start.degree = 90,
    track.margin = c(0.01, 0.01),
    cell.padding = c(0, 0, 0, 0)
  )

  circos.initialize(
    factors = nodes,
    xlim = cbind(rep(0, length(nodes)), rep(1, length(nodes)))
  )

  if (with_labels) {
    circos.trackPlotRegion(
      ylim = c(0, 1),
      track.height = 0.08,
      bg.border = NA,
      panel.fun = function(x, y) {
        sector <- get.cell.meta.data("sector.index")
        circos.text(
          x = 0.5,
          y = 2.6,
          labels = shorten_celltype(sector),
          facing = "bending.outside",
          niceFacing = TRUE,
          cex = label_cex
        )
      }
    )
  } else {
    circos.trackPlotRegion(
      ylim = c(0, 1),
      track.height = 0.001,
      bg.border = NA,
      panel.fun = function(x, y) {}
    )
  }
}

draw_nonself_links <- function(df_nonself) {
  if (nrow(df_nonself) == 0) {
    return(invisible(NULL))
  }

  for (i in seq_len(nrow(df_nonself))) {
    lw <- df_nonself$link_width[i]
    circos.link(
      df_nonself$Sender[i], 0.5,
      df_nonself$Receiver[i], 0.5,
      col = df_nonself$link_col[i],
      border = NA,
      lwd = lw,
      directional = 1,
      arr.length = 0.3,
      arr.width = 0.3,
      arr.lwd = lw * 0.4
    )
  }
}

draw_self_loops <- function(df_self) {
  if (nrow(df_self) == 0) {
    return(invisible(NULL))
  }

  df_self <- df_self %>%
    group_by(Sender) %>%
    mutate(
      loop_id = row_number(),
      loop_n  = n()
    ) %>%
    ungroup()

  for (i in seq_len(nrow(df_self))) {
    nd  <- df_self$Sender[i]
    lw  <- df_self$link_width[i]
    col <- df_self$link_col[i]

    # Spread multiple self-loops slightly within the node sector.
    t <- if (df_self$loop_n[i] == 1) {
      0
    } else {
      (df_self$loop_id[i] - 1) / (df_self$loop_n[i] - 1)
    }
    center <- 0.5 + (t - 0.5) * 0.08

    # Increase the loop span slightly for stronger links.
    span <- 0.055 + 0.015 * scales::rescale(lw, to = c(0, 1))
    span <- 4 * span
    x1 <- max(0.02, min(0.98, center - span))
    x2 <- max(0.02, min(0.98, center + span))

    # Keep self-loops close to the node ring.
    rou <- 0.97
    hgt <- 0.045 + 0.03 * scales::rescale(lw, to = c(0, 1))

    circos.link(
      nd, x1,
      nd, x2,
      col = col,
      border = NA,
      lwd = lw,
      directional = 1,
      arr.length = 0.22,
      arr.width = 0.22,
      arr.lwd = lw * 0.4,
      rou1 = rou,
      rou2 = rou,
      h = -hgt
    )
  }
}

draw_node_points <- function(nodes, celltype_colors, point_cex = 3.2) {
  for (nd in nodes) {
    circos.points(
      x = 0.5,
      y = 0,
      sector.index = nd,
      pch = 21,
      cex = point_cex,
      bg = celltype_colors[nd],
      col = "black"
    )
  }
}

draw_labeled_circos <- function(df_plot, nodes, celltype_colors, output_pdf) {
  pdf(output_pdf, width = 10, height = 10)

  initialize_circos_layout(nodes = nodes, with_labels = TRUE, label_cex = 1.4)

  df_self    <- df_plot %>% filter(Sender == Receiver)
  df_nonself <- df_plot %>% filter(Sender != Receiver)

  draw_nonself_links(df_nonself)
  draw_self_loops(df_self)
  draw_node_points(nodes, celltype_colors, point_cex = 3.2)

  circos.clear()
  dev.off()
}

draw_nolabel_circos <- function(df_plot, nodes, celltype_colors, output_png) {
  png(output_png, width = 1800, height = 1800, res = 300)

  initialize_circos_layout(nodes = nodes, with_labels = FALSE)
  draw_nonself_links(df_plot)
  draw_node_points(nodes, celltype_colors, point_cex = 2.2)

  circos.clear()
  dev.off()
}
```

# 4. Read input data

```{r read-data}
start_time <- Sys.time()

if (!file.exists(mi_link_file)) {
  stop(paste("Cannot find input file:", mi_link_file))
}

mi_links <- read_pancancer_links(mi_link_file)

cat("Loaded sender-receiver links:", nrow(mi_links), "\n")
print(head(mi_links))
```

# 5. Prepare node colors and plotting dataframe

```{r prepare-data}
celltype_colors <- setNames(color_set[seq_along(celltypes)], celltypes)

nodes <- get_nodes_in_order(mi_links, celltypes)
if (length(nodes) == 0) {
  stop("None of the requested cell types are present in the MI link table.")
}

df_plot <- prepare_link_dataframe(
  df = mi_links,
  link_power = link_power,
  link_width_range = link_width_range,
  quantile_floor_hi = quantile_floor_hi,
  quantile_floor_lo = quantile_floor_lo
)

cat("Nodes shown on the circle:", length(nodes), "\n")
print(nodes)
```

# 6. Draw the labeled circos plot

This version includes sector labels and is typically used as the main figure.

```{r draw-labeled-circos}
draw_labeled_circos(
  df_plot = df_plot,
  nodes = nodes,
  celltype_colors = celltype_colors,
  output_pdf = output_pdf_labeled
)

cat("Saved labeled plot to:", output_pdf_labeled, "\n")
```

# 7. Draw the no-label circos plot

This version omits sector labels and is useful when labels are added later in Illustrator.

```{r draw-nolabel-circos}
draw_nolabel_circos(
  df_plot = df_plot,
  nodes = nodes,
  celltype_colors = celltype_colors,
  output_png = output_png_nolabel
)

cat("Saved no-label plot to:", output_png_nolabel, "\n")
```

# 8. Session summary

```{r summary}
end_time <- Sys.time()
cat("Finished.\n")
cat("Elapsed time:", round(difftime(end_time, start_time, units = "secs"), 2), "seconds\n")
```

# Notes for adapting this notebook

When adapting this notebook to another MI file or another dimension, the sections that usually need to change are:

- `file_savepath_main`, input file name, and output file names;
- the cell-type order in `celltypes`;
- the color palette in `color_set`;
- the link-weight transform and rescaling settings;
- whether the no-label output should be PDF or PNG.

The helper functions for reading, preprocessing, node ordering, and circos plotting can usually remain unchanged as long as the input table contains the columns `Sender`, `Receiver`, and `Mean`.
