---
title: "GO Enrichment Visualization for MI2"
author: "Junjie Tang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# Overview

This R Markdown document generates two final GO enrichment bubble plots for MI2:

- **p1**: GO enrichment of top regulators (sender side)
- **p2**: GO enrichment of top targets (receiver side)

The workflow is organized into four steps:

1. Load required packages and input files.
2. Clean and standardize the GO enrichment tables.
3. Define the final term order for the sender and receiver plots.
4. Generate and save the two final figures.

Only the final figures `p1` and `p2` are kept in this version. Intermediate figures from the original script have been removed.

# Load packages and define paths

```{r packages-and-paths}
library(ggplot2)
library(dplyr)
library(stringr)

result_dir <- "D:/SpiderNet/Results/Pancancer/V1/SpiderNet_Result_dim11"

toptarget_file <- file.path(result_dir, "GO_enrichment_toptarget_MI2.csv")
topregulator_file <- file.path(result_dir, "GO_enrichment_topregulator_MI2.csv")
```

# Read input tables

```{r read-data}
GO_enrichment_toptarget_MI2 <- read.csv(toptarget_file)
GO_enrichment_topregulator_MI2 <- read.csv(topregulator_file)
GO_enrichment_toptarget_MI2
GO_enrichment_topregulator_MI2
```

# Define helper function for data cleaning

```{r helper-function}
prepare_go_df <- function(df) {
  df %>%
    mutate(
      # Remove GO accession IDs such as "(GO:XXXXXXX)" from the term labels.
      Term_clean = str_replace(Term, "\\s*\\(GO:\\d+\\)", ""),
      # Extract the number of overlapping genes from strings such as "5/120".
      GeneCount = as.numeric(sub("/.*", "", Overlap)),
      # Convert adjusted p-values to -log10 scale for plotting.
      logP = -log10(Adjusted.P.value)
    )
}
```

# Prepare sender and receiver enrichment tables

```{r prepare-data}
df_sender <- prepare_go_df(GO_enrichment_topregulator_MI2)
df_receiver <- prepare_go_df(GO_enrichment_toptarget_MI2)

# Define the final display order for the sender enrichment plot (p1).
order_p1 <- rev(c(
    # [1] Inflammatory / cytokine output axis (Sender)
    "positive regulation of interleukin-6 production",
    "positive regulation of interleukin-8 production",
    "positive regulation of chemokine production",

    # [2] Growth signal / kinase activation axis (Sender)
    "regulation of ERBB signaling pathway",
    "positive regulation of protein kinase B signaling",
    "positive regulation of MAPK cascade"
))

# Define the final display order for the receiver enrichment plot (p2).
order_p2 <- rev(c(
    # [1-1] Inflammatory / STAT response program (Receiver)
    "cellular response to cytokine stimulus",
    "regulation of receptor signaling pathway via STAT",

    # [1-2] Inflammation-driven remodeling program (Receiver)
    "extracellular matrix organization",
    
    # [2] Cell-cycle / proliferative execution program (Receiver)
    "DNA replication",
    "G2/M transition of mitotic cell cycle"
))

df_sender <- df_sender %>%
  filter(Term_clean %in% order_p1) %>%
  mutate(Term_clean = factor(Term_clean, levels = order_p1))

df_receiver <- df_receiver %>%
  filter(Term_clean %in% order_p2) %>%
  mutate(Term_clean = factor(Term_clean, levels = order_p2))
```

# Define shared plotting scale

```{r shared-scale}
# Use a shared point-size scale so that p1 and p2 are visually comparable.
all_logP <- c(df_sender$logP, df_receiver$logP)
min_logP <- min(all_logP)
max_logP <- max(all_logP)

size_breaks <- c(2, 5, 8)
```

# Plot p1: sender-side GO enrichment

```{r plot-p1, fig.width=7.8, fig.height=7}
p1 <- ggplot(df_sender, aes(x = 1, y = Term_clean)) +
  geom_point(aes(size = logP), color = "#FD2D2D") +
  scale_size(
    range = c(7, 15),
    limits = c(min_logP, max_logP),
    breaks = size_breaks,
    name = "-log10(P)"
  ) +
  scale_y_discrete(position = "left") +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 27, lineheight = 0.9)
  )

p1
```

```{r save-p1, eval=TRUE}
ggsave(
  file.path(result_dir, "GO_enrichment_topregulator_MI2_SingleColumn_BubblePlot.png"),
  plot = p1,
  width = 10,
  height = 7,
  dpi = 300
)

ggsave(
  file.path(result_dir, "GO_enrichment_topregulator_MI2_SingleColumn_BubblePlot.pdf"),
  plot = p1,
  width = 10,
  height = 7,
  dpi = 300
)
```

# Plot p2: receiver-side GO enrichment

```{r plot-p2, fig.width=12, fig.height=7}
p2 <- ggplot(df_receiver, aes(x = 1, y = Term_clean)) +
  geom_point(aes(size = logP), color = "#FD2D2D") +
  scale_size(
    range = c(7, 15),
    limits = c(min_logP, max_logP),
    breaks = size_breaks,
    name = "-log10(P)"
  ) +
  scale_y_discrete(position = "right") +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 27, lineheight = 0.9)
  )

p2
```

```{r save-p2, eval=TRUE}
ggsave(
  file.path(result_dir, "GO_enrichment_toptarget_MI2_SingleColumn_BubblePlot.png"),
  plot = p2,
  width = 12,
  height = 7,
  dpi = 300
)

ggsave(
  file.path(result_dir, "GO_enrichment_toptarget_MI2_SingleColumn_BubblePlot.pdf"),
  plot = p2,
  width = 12,
  height = 7,
  dpi = 300
)
```

# Session information

```{r session-info}
sessionInfo()
```
