---
title: "MI–CAF and Functional State Boxplot Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Overview

This notebook generates boxplots to compare CAF scores or malignant functional-state scores between cells with **high** versus **low** MI levels.

The workflow includes:

1. Loading precomputed CSV files for each MI–phenotype comparison.
2. Automatically detecting the relevant x-axis and y-axis columns.
3. Dividing cells into two groups using a user-defined MI threshold.
4. Performing a Wilcoxon rank-sum test between the two groups.
5. Saving each boxplot as an individual PDF file.
6. Summarizing the statistical results in a single table for easy review.
7. Displaying all generated boxplots at the end of the notebook.

This notebook is organized so that users can quickly identify what needs to be modified when applying the workflow to a new dataset.

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(knitr)
  library(tibble)
})
```

# 2. User settings

Modify this section first when applying the notebook to a new result folder.

```{r user-settings}
# =========================================================
# Main input and output settings
# =========================================================
file_savepath_main <- "D:/SpiderNet/Results/HGSOC/V1/SpiderNet_Result_dim15/"

# Threshold used to split MI level into high and low groups
thr_use <- 0.5
thr_tag <- gsub("\\.", "p", sprintf("%.2f", thr_use))

# =========================================================
# Candidate column names
# The script will automatically use the first matching column
# =========================================================
x_candidates <- c(
  "MI_Level",
  "Sending_MI10", "Sending_MI6", "Sending_MI13",
  "Receiving_MI10", "Receiving_MI6", "Receiving_MI13"
)

caf_y_candidates <- c("CAF_Score", "CAF_score")
state_y_candidates <- c("State_Score", "state_score")

# =========================================================
# Plot colors
# =========================================================
caf_fill_low    <- "#C2E2FA"
caf_fill_high   <- "#B7A3E3"
state_fill_low  <- "#FFF1CB"
state_fill_high <- "#FF8F8F"
```

# 3. Helper functions

These functions handle significance labels, automatic column selection, PDF export, summary statistics, and boxplot generation.

```{r helper-functions}
# =========================================================
# Convert p-values to significance labels
# =========================================================
p_to_star <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-4) return("****")
  if (p < 1e-3) return("***")
  if (p < 1e-2) return("**")
  if (p < 0.05) return("*")
  return("ns")
}

# =========================================================
# Return the first existing column from a candidate list
# =========================================================
pick_first_existing <- function(df, candidates) {
  for (nm in candidates) {
    if (nm %in% colnames(df)) return(nm)
  }
  stop(
    paste0(
      "None of these columns exist: ",
      paste(candidates, collapse = ", ")
    )
  )
}

# =========================================================
# Save a plot as a PDF file
# =========================================================
save_plot_pdf <- function(p, fn, width = 1.2, height = 2.6, out_dir = file_savepath_main) {
  full <- file.path(out_dir, fn)
  has_cairo <- capabilities("cairo")

  if (has_cairo) {
    ggsave(
      filename = full,
      plot = p,
      width = width,
      height = height,
      dpi = 300,
      device = cairo_pdf,
      bg = "white"
    )
  } else {
    ggsave(
      filename = full,
      plot = p,
      width = width,
      height = height,
      dpi = 300,
      bg = "white"
    )
  }
}

# =========================================================
# Define a compact Nature-style plotting theme
# =========================================================
theme_nature <- function(base_size = 9, base_family = "Arial") {
  theme_classic(base_size = base_size, base_family = base_family) %+replace%
    theme(
      plot.title = element_text(size = base_size, face = "plain", hjust = 0, color = "black"),
      axis.title = element_text(size = base_size, color = "black"),
      axis.text  = element_text(size = base_size - 1, color = "black"),
      axis.line  = element_line(color = "black", linewidth = 0.35),
      axis.ticks = element_line(color = "black", linewidth = 0.35),
      axis.ticks.length = grid::unit(2, "pt"),
      panel.border = element_blank(),
      panel.grid = element_blank(),
      legend.position = "none",
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}

# =========================================================
# Summarize two groups defined by an MI threshold
# =========================================================
summarize_two_groups <- function(df, xcol, ycol, thr = 0.25) {
  lvl_low  <- paste0("<", thr)
  lvl_high <- paste0("\u2265", thr)

  df2 <- df %>%
    dplyr::filter(is.finite(.data[[xcol]]), is.finite(.data[[ycol]])) %>%
    dplyr::mutate(
      MI_group = ifelse(.data[[xcol]] >= thr, lvl_high, lvl_low),
      MI_group = factor(MI_group, levels = c(lvl_high, lvl_low))
    )

  n_high <- sum(df2$MI_group == lvl_high, na.rm = TRUE)
  n_low  <- sum(df2$MI_group == lvl_low, na.rm = TRUE)

  m_high <- mean(df2[[ycol]][df2$MI_group == lvl_high], na.rm = TRUE)
  m_low  <- mean(df2[[ycol]][df2$MI_group == lvl_low], na.rm = TRUE)
  mean_diff <- m_high - m_low

  if (is.finite(m_high) && is.finite(m_low) && m_high > 0 && m_low > 0) {
    log2_mean_change <- log2(m_high / m_low)
  } else {
    log2_mean_change <- NA_real_
  }

  p_wilcox <- NA_real_
  if (length(unique(stats::na.omit(df2$MI_group))) == 2) {
    wt <- suppressWarnings(
      wilcox.test(df2[[ycol]] ~ df2$MI_group, alternative = "two.sided")
    )
    p_wilcox <- wt$p.value
  }

  tibble(
    x_column = xcol,
    y_column = ycol,
    threshold = thr,
    n_high = n_high,
    n_low = n_low,
    mean_high = m_high,
    mean_low = m_low,
    mean_diff = mean_diff,
    log2_mean_change = log2_mean_change,
    p_wilcox = p_wilcox,
    significance = p_to_star(p_wilcox)
  )
}

# =========================================================
# Build a boxplot comparing high and low MI groups
# Notes:
# - The high group (>= threshold) is shown first
# - The low group (< threshold) is shown second
# - Whisker caps are explicitly drawn with stat_boxplot()
# =========================================================
make_box_plot <- function(
    df,
    xcol,
    ycol,
    thr = 0.25,
    fill_low = "#C2E2FA",
    fill_high = "#B7A3E3",
    title_txt = NULL,
    xlab = NULL,
    ylab = NULL
) {
  lvl_low  <- paste0("<", thr)
  lvl_high <- paste0("\u2265", thr)

  df2 <- df %>%
    dplyr::filter(is.finite(.data[[xcol]]), is.finite(.data[[ycol]])) %>%
    dplyr::mutate(
      MI_group = ifelse(.data[[xcol]] >= thr, lvl_high, lvl_low),
      MI_group = factor(MI_group, levels = c(lvl_high, lvl_low))
    )

  stats_df <- summarize_two_groups(df2, xcol, ycol, thr = thr)
  p_wilcox <- stats_df$p_wilcox[[1]]
  star <- stats_df$significance[[1]]

  p_lbl <- ifelse(
    is.na(p_wilcox),
    "p = NA",
    paste0("p = ", format(p_wilcox, digits = 2, scientific = TRUE))
  )

  y_top <- max(df2[[ycol]], na.rm = TRUE)
  y_min <- min(df2[[ycol]], na.rm = TRUE)
  y_rng <- y_top - y_min + 1e-12
  y_anno <- y_top + 0.15 * y_rng

  pal <- setNames(c(fill_high, fill_low), c(lvl_high, lvl_low))

  ggplot(df2, aes(x = MI_group, y = .data[[ycol]], fill = MI_group)) +
    stat_boxplot(
      geom = "errorbar",
      width = 0.22,
      linewidth = 0.35,
      color = "black"
    ) +
    geom_boxplot(
      width = 0.58,
      outlier.shape = NA,
      linewidth = 0.35,
      color = "black"
    ) +
    scale_fill_manual(values = pal, drop = FALSE) +
    annotate(
      "segment",
      x = 1, xend = 2,
      y = y_anno, yend = y_anno,
      linewidth = 0.35, color = "black"
    ) +
    annotate(
      "segment",
      x = 1, xend = 1,
      y = y_anno, yend = y_anno - 0.05 * y_anno,
      linewidth = 0.35, color = "black"
    ) +
    annotate(
      "segment",
      x = 2, xend = 2,
      y = y_anno, yend = y_anno - 0.05 * y_anno,
      linewidth = 0.35, color = "black"
    ) +
    annotate(
      "text",
      x = 1.5,
      y = y_anno + 0.03 * y_anno,
      label = paste0(p_lbl, "  ", star),
      size = 3.0,
      vjust = 0,
      family = "Arial",
      color = "black"
    ) +
    coord_cartesian(clip = "off") +
    labs(title = title_txt, x = xlab, y = ylab) +
    theme_nature(base_size = 9, base_family = "Arial") +
    theme(axis.text.x = element_text(size = 8, color = "black"))
}
```

# 4. Load input data

Each CSV file corresponds to one MI–phenotype comparison.

```{r load-data}
data_list <- list(
  CAF_MI10           = read.csv(file.path(file_savepath_main, "Malignant_COI5_CAF_score_vs_Sending_MI10_scatter.csv")),
  CAF_MI6            = read.csv(file.path(file_savepath_main, "Malignant_COI5_CAF_score_vs_Sending_MI6_scatter.csv")),
  CAF_MI13           = read.csv(file.path(file_savepath_main, "Malignant_COI5_CAF_score_vs_Sending_MI13_scatter.csv")),

  Metastasis_MI10    = read.csv(file.path(file_savepath_main, "Malignant_COI5_Metastasis_vs_Receiving_MI10_scatter.csv")),
  Metastasis_MI6     = read.csv(file.path(file_savepath_main, "Malignant_COI5_Metastasis_vs_Receiving_MI6_scatter.csv")),
  Metastasis_MI13    = read.csv(file.path(file_savepath_main, "Malignant_COI5_Metastasis_vs_Receiving_MI13_scatter.csv")),

  Hypoxia_MI10       = read.csv(file.path(file_savepath_main, "Malignant_COI5_Hypoxia_vs_Receiving_MI10_scatter.csv")),
  Hypoxia_MI6        = read.csv(file.path(file_savepath_main, "Malignant_COI5_Hypoxia_vs_Receiving_MI6_scatter.csv")),
  Hypoxia_MI13       = read.csv(file.path(file_savepath_main, "Malignant_COI5_Hypoxia_vs_Receiving_MI13_scatter.csv")),

  Angiogenesis_MI10  = read.csv(file.path(file_savepath_main, "Malignant_COI5_Angiogenesis_vs_Receiving_MI10_scatter.csv")),
  Angiogenesis_MI6   = read.csv(file.path(file_savepath_main, "Malignant_COI5_Angiogenesis_vs_Receiving_MI6_scatter.csv")),
  Angiogenesis_MI13  = read.csv(file.path(file_savepath_main, "Malignant_COI5_Angiogenesis_vs_Receiving_MI13_scatter.csv")),

  Inflammation_MI6   = read.csv(file.path(file_savepath_main, "Malignant_COI5_Inflammation_vs_Receiving_MI6_scatter.csv")),
  Inflammation_MI10  = read.csv(file.path(file_savepath_main, "Malignant_COI5_Inflammation_vs_Receiving_MI10_scatter.csv")),
  Inflammation_MI13  = read.csv(file.path(file_savepath_main, "Malignant_COI5_Inflammation_vs_Receiving_MI13_scatter.csv"))
)

names(data_list)
```

# 5. Define the analysis plan

This table specifies which dataset to use, how the output file should be named, which candidate y-axis columns should be checked, and which color scheme should be applied.

```{r analysis-specs}
analysis_specs <- list(
  list(key = "CAF_MI10",          prefix = "CAF_vs_SendingMI10",            y_candidates = caf_y_candidates,   fill_low = caf_fill_low,   fill_high = caf_fill_high),
  list(key = "CAF_MI6",           prefix = "CAF_vs_SendingMI6",             y_candidates = caf_y_candidates,   fill_low = caf_fill_low,   fill_high = caf_fill_high),
  list(key = "CAF_MI13",          prefix = "CAF_vs_SendingMI13",            y_candidates = caf_y_candidates,   fill_low = caf_fill_low,   fill_high = caf_fill_high),

  list(key = "Metastasis_MI10",   prefix = "Metastasis_vs_ReceivingMI10",   y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Metastasis_MI6",    prefix = "Metastasis_vs_ReceivingMI6",    y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Metastasis_MI13",   prefix = "Metastasis_vs_ReceivingMI13",   y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),

  list(key = "Hypoxia_MI10",      prefix = "Hypoxia_vs_ReceivingMI10",      y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Hypoxia_MI6",       prefix = "Hypoxia_vs_ReceivingMI6",       y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Hypoxia_MI13",      prefix = "Hypoxia_vs_ReceivingMI13",      y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),

  list(key = "Angiogenesis_MI10", prefix = "Angiogenesis_vs_ReceivingMI10", y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Angiogenesis_MI6",  prefix = "Angiogenesis_vs_ReceivingMI6",  y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Angiogenesis_MI13", prefix = "Angiogenesis_vs_ReceivingMI13", y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),

  list(key = "Inflammation_MI6",  prefix = "Inflammation_vs_ReceivingMI6",  y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Inflammation_MI10", prefix = "Inflammation_vs_ReceivingMI10", y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Inflammation_MI13", prefix = "Inflammation_vs_ReceivingMI13", y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high)
)
```

# 6. Run the analysis and save boxplots

For each comparison, the notebook will detect the relevant columns, compute summary statistics, generate the boxplot, save it as a PDF, and store the result for later review.

```{r run-analysis}
plot_results <- list()
all_boxplots <- list()
summary_list <- list()

for (spec in analysis_specs) {
  message("Running: ", spec$prefix)

  df <- data_list[[spec$key]]

  xcol <- pick_first_existing(df, x_candidates)
  ycol <- pick_first_existing(df, spec$y_candidates)

  stats_df <- summarize_two_groups(
    df = df,
    xcol = xcol,
    ycol = ycol,
    thr = thr_use
  ) %>%
    mutate(
      comparison = spec$prefix,
      .before = 1
    )

  p_box <- make_box_plot(
    df = df,
    xcol = xcol,
    ycol = ycol,
    thr = thr_use,
    fill_low = spec$fill_low,
    fill_high = spec$fill_high,
    title_txt = NULL,
    xlab = NULL,
    ylab = NULL
  )

  save_plot_pdf(
    p = p_box,
    fn = paste0(spec$prefix, "_box_wilcox_thr", thr_tag, ".pdf"),
    width = 1.2,
    height = 2.6
  )

  plot_results[[spec$key]] <- list(
    data = df,
    xcol = xcol,
    ycol = ycol,
    stats = stats_df,
    box = p_box
  )

  all_boxplots[[spec$prefix]] <- p_box
  summary_list[[spec$prefix]] <- stats_df
}
```

# 7. Summary table

This table provides a compact overview of all comparisons.

```{r summary-table}
summary_df <- bind_rows(summary_list) %>%
  mutate(
    mean_high = signif(mean_high, 4),
    mean_low = signif(mean_low, 4),
    mean_diff = signif(mean_diff, 4),
    log2_mean_change = signif(log2_mean_change, 4),
    p_wilcox = signif(p_wilcox, 4)
  )

kable(summary_df, align = "lcccccccccc")
```

# 8. Show all boxplots

This section prints all boxplots for quick inspection inside the notebook.

```{r show-plots, fig.width=2.2, fig.height=3.2}
for (nm in names(all_boxplots)) {
  message("Showing plot: ", nm)
  print(all_boxplots[[nm]])
}
```

# 9. Notes for users

## What usually needs to be changed

In most cases, users only need to modify:

- `file_savepath_main`
- `thr_use`
- the file names in `data_list`
- the analysis mapping in `analysis_specs`

## Expected input format

Each CSV file should contain:

- one MI-related column, such as `MI_Level`, `Sending_MI10`, or `Receiving_MI10`
- one phenotype score column, such as `CAF_Score` or `State_Score`

## Output files

The notebook saves one PDF file per comparison to the following directory:

```{r output-path, echo=FALSE}
cat(file_savepath_main)
```

The output file names follow this pattern:

- `CAF_vs_SendingMI10_box_wilcox_thr0p50.pdf`
- `Metastasis_vs_ReceivingMI10_box_wilcox_thr0p50.pdf`

# Session information

```{r session-info}
sessionInfo()
```
