---
title: "MI–CAF and Functional State Boxplot Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Overview

Plot the saved CAF and malignant-program scores with the original R Wilcoxon tests and SMD annotations.

Run `python run_benchmarks.py --stage r-plots --plot-only` from this folder, or render this Rmd with R 4.6.1, `rmarkdown`, `knitr`, `ggplot2`, `dplyr`, `tibble`, and Pandoc.
Inputs resolve from `HGSOC_LOCAL_RUN` (default `output/V1/SpiderNet_Result_dim15`) then `HGSOC_RUN_DIR` (default the matching shared `Results/HGSOC` run).
PDF/PNG/CSV outputs use `HGSOC_R_OUTPUT` (default `output/r-plots`); the runner saves the executed document under `output/executed`.
The R and Python statistical variants remain separate. Thresholds, tests, colors, and program selections are unchanged.

# 1. Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(knitr)
  library(tibble)
})
```

# 2. User settings

Set the input/output environment variables above when relocating data. Keep the recorded scientific settings unchanged for reproduction.

```{r user-settings}
# =========================================================
# Main input and output settings
# =========================================================
# Run from the HGSOC folder; the Python runner supplies these same paths.
local_run_R <- Sys.getenv(
  "HGSOC_LOCAL_RUN", file.path(getwd(), "output", "V1", "SpiderNet_Result_dim15")
)
upstream_run_R <- Sys.getenv(
  "HGSOC_RUN_DIR", "D:/SpiderNet/Results/HGSOC/V1/SpiderNet_Result_dim15"
)
file_savepath_main <- Sys.getenv(
  "HGSOC_R_OUTPUT", file.path(getwd(), "output", "r-plots")
)
dir.create(file_savepath_main, recursive = TRUE, showWarnings = FALSE)

input_path_R <- function(filename) {
  local_path <- file.path(local_run_R, filename)
  if (file.exists(local_path)) return(local_path)
  upstream_path <- file.path(upstream_run_R, filename)
  if (file.exists(upstream_path)) return(upstream_path)
  local_path
}

# Threshold used to split MI level into high and low groups
thr_use <- 0.5
thr_tag <- gsub("\\.", "p", sprintf("%.2f", thr_use))

# =========================================================
# Candidate column names
# The script will automatically use the first matching column
# =========================================================
x_candidates <- c(
  "MI_Level",
  "Sending_MI10", "Sending_MI6", "Sending_MI13",
  "Receiving_MI10", "Receiving_MI6", "Receiving_MI13"
)

caf_y_candidates <- c("CAF_Score", "CAF_score")
state_y_candidates <- c("State_Score", "state_score")

# =========================================================
# Plot colors
# =========================================================
caf_fill_low    <- "#C2E2FA"
caf_fill_high   <- "#B7A3E3"
state_fill_low  <- "#FFF1CB"
state_fill_high <- "#FF8F8F"
```

# 3. Helper functions

These functions handle significance labels, automatic column selection, PDF export, summary statistics, and boxplot generation.

```{r helper-functions}
# =========================================================
# Convert p-values to significance labels
# =========================================================
p_to_star <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-4) return("****")
  if (p < 1e-3) return("***")
  if (p < 1e-2) return("**")
  if (p < 0.05) return("*")
  return("ns")
}

# =========================================================
# Return the first existing column from a candidate list
# =========================================================
pick_first_existing <- function(df, candidates) {
  for (nm in candidates) {
    if (nm %in% colnames(df)) return(nm)
  }
  stop(
    paste0(
      "None of these columns exist: ",
      paste(candidates, collapse = ", ")
    )
  )
}

# =========================================================
# Save a plot as a PDF file
# =========================================================
save_plot_pdf <- function(p, fn, width = 1.2, height = 2.6, out_dir = file_savepath_main) {
  full <- file.path(out_dir, fn)
  has_cairo <- capabilities("cairo")

  if (has_cairo) {
    ggsave(
      filename = full,
      plot = p,
      width = width,
      height = height,
      dpi = 300,
      device = cairo_pdf,
      bg = "white"
    )
  } else {
    ggsave(
      filename = full,
      plot = p,
      width = width,
      height = height,
      dpi = 300,
      bg = "white"
    )
  }
}

# =========================================================
# Define a compact Nature-style plotting theme
# =========================================================
theme_nature <- function(base_size = 9, base_family = "Arial") {
  theme_classic(base_size = base_size, base_family = base_family) %+replace%
    theme(
      plot.title = element_text(size = base_size, face = "plain", hjust = 0, color = "black"),
      axis.title = element_text(size = base_size, color = "black"),
      axis.text  = element_text(size = base_size - 1, color = "black"),
      axis.line  = element_line(color = "black", linewidth = 0.35),
      axis.ticks = element_line(color = "black", linewidth = 0.35),
      axis.ticks.length = grid::unit(2, "pt"),
      panel.border = element_blank(),
      panel.grid = element_blank(),
      legend.position = "none",
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}

# =========================================================
# Summarize two groups defined by an MI threshold
# =========================================================
summarize_two_groups <- function(df, xcol, ycol, thr = 0.25) {
  lvl_low  <- paste0("<", thr)
  lvl_high <- paste0("\u2265", thr)

  df2 <- df %>%
    dplyr::filter(is.finite(.data[[xcol]]), is.finite(.data[[ycol]])) %>%
    dplyr::mutate(
      MI_group = ifelse(.data[[xcol]] >= thr, lvl_high, lvl_low),
      MI_group = factor(MI_group, levels = c(lvl_high, lvl_low))
    )

  n_high <- sum(df2$MI_group == lvl_high, na.rm = TRUE)
  n_low  <- sum(df2$MI_group == lvl_low, na.rm = TRUE)

  m_high <- mean(df2[[ycol]][df2$MI_group == lvl_high], na.rm = TRUE)
  m_low  <- mean(df2[[ycol]][df2$MI_group == lvl_low], na.rm = TRUE)
  mean_diff <- m_high - m_low

  if (is.finite(m_high) && is.finite(m_low) && m_high > 0 && m_low > 0) {
    log2_mean_change <- log2(m_high / m_low)
  } else {
    log2_mean_change <- NA_real_
  }

  p_wilcox <- NA_real_
  if (length(unique(stats::na.omit(df2$MI_group))) == 2) {
    wt <- suppressWarnings(
      wilcox.test(df2[[ycol]] ~ df2$MI_group, alternative = "two.sided")
    )
    p_wilcox <- wt$p.value
  }

  tibble(
    x_column = xcol,
    y_column = ycol,
    threshold = thr,
    n_high = n_high,
    n_low = n_low,
    mean_high = m_high,
    mean_low = m_low,
    mean_diff = mean_diff,
    log2_mean_change = log2_mean_change,
    p_wilcox = p_wilcox,
    significance = p_to_star(p_wilcox)
  )
}

# =========================================================
# Build a boxplot comparing high and low MI groups
# Notes:
# - The high group (>= threshold) is shown first
# - The low group (< threshold) is shown second
# - Whisker caps are explicitly drawn with stat_boxplot()
# =========================================================
make_box_plot <- function(
    df,
    xcol,
    ycol,
    thr = 0.25,
    fill_low = "#C2E2FA",
    fill_high = "#B7A3E3",
    title_txt = NULL,
    xlab = NULL,
    ylab = NULL
) {
  lvl_low  <- paste0("<", thr)
  lvl_high <- paste0("\u2265", thr)

  df2 <- df %>%
    dplyr::filter(is.finite(.data[[xcol]]), is.finite(.data[[ycol]])) %>%
    dplyr::mutate(
      MI_group = ifelse(.data[[xcol]] >= thr, lvl_high, lvl_low),
      MI_group = factor(MI_group, levels = c(lvl_high, lvl_low))
    )

  stats_df <- summarize_two_groups(df2, xcol, ycol, thr = thr)
  p_wilcox <- stats_df$p_wilcox[[1]]
  star <- stats_df$significance[[1]]

  p_lbl <- ifelse(
    is.na(p_wilcox),
    "p = NA",
    paste0("p = ", format(p_wilcox, digits = 2, scientific = TRUE))
  )

  y_top <- max(df2[[ycol]], na.rm = TRUE)
  y_min <- min(df2[[ycol]], na.rm = TRUE)
  y_rng <- y_top - y_min + 1e-12
  y_anno <- y_top + 0.15 * y_rng

  pal <- setNames(c(fill_high, fill_low), c(lvl_high, lvl_low))

  ggplot(df2, aes(x = MI_group, y = .data[[ycol]], fill = MI_group)) +
    stat_boxplot(
      geom = "errorbar",
      width = 0.22,
      linewidth = 0.35,
      color = "black"
    ) +
    geom_boxplot(
      width = 0.58,
      outlier.shape = NA,
      linewidth = 0.35,
      color = "black"
    ) +
    scale_fill_manual(values = pal, drop = FALSE) +
    annotate(
      "segment",
      x = 1, xend = 2,
      y = y_anno, yend = y_anno,
      linewidth = 0.35, color = "black"
    ) +
    annotate(
      "segment",
      x = 1, xend = 1,
      y = y_anno, yend = y_anno - 0.05 * y_anno,
      linewidth = 0.35, color = "black"
    ) +
    annotate(
      "segment",
      x = 2, xend = 2,
      y = y_anno, yend = y_anno - 0.05 * y_anno,
      linewidth = 0.35, color = "black"
    ) +
    annotate(
      "text",
      x = 1.5,
      y = y_anno + 0.03 * y_anno,
      label = paste0(p_lbl, "  ", star),
      size = 3.0,
      vjust = 0,
      family = "Arial",
      color = "black"
    ) +
    coord_cartesian(clip = "off") +
    labs(title = title_txt, x = xlab, y = ylab) +
    theme_nature(base_size = 9, base_family = "Arial") +
    theme(axis.text.x = element_text(size = 8, color = "black"))
}
```

# 4. Load input data

Each CSV file corresponds to one MI–phenotype comparison.

```{r load-data}
data_list <- list(
  CAF_MI10           = read.csv(input_path_R("Malignant_COI5_CAF_score_vs_Sending_MI10_scatter.csv")),
  CAF_MI6            = read.csv(input_path_R("Malignant_COI5_CAF_score_vs_Sending_MI6_scatter.csv")),
  CAF_MI13           = read.csv(input_path_R("Malignant_COI5_CAF_score_vs_Sending_MI13_scatter.csv")),

  Metastasis_MI10    = read.csv(input_path_R("Malignant_COI5_Metastasis_vs_Receiving_MI10_scatter.csv")),
  Metastasis_MI6     = read.csv(input_path_R("Malignant_COI5_Metastasis_vs_Receiving_MI6_scatter.csv")),
  Metastasis_MI13    = read.csv(input_path_R("Malignant_COI5_Metastasis_vs_Receiving_MI13_scatter.csv")),

  Hypoxia_MI10       = read.csv(input_path_R("Malignant_COI5_Hypoxia_vs_Receiving_MI10_scatter.csv")),
  Hypoxia_MI6        = read.csv(input_path_R("Malignant_COI5_Hypoxia_vs_Receiving_MI6_scatter.csv")),
  Hypoxia_MI13       = read.csv(input_path_R("Malignant_COI5_Hypoxia_vs_Receiving_MI13_scatter.csv")),

  Angiogenesis_MI10  = read.csv(input_path_R("Malignant_COI5_Angiogenesis_vs_Receiving_MI10_scatter.csv")),
  Angiogenesis_MI6   = read.csv(input_path_R("Malignant_COI5_Angiogenesis_vs_Receiving_MI6_scatter.csv")),
  Angiogenesis_MI13  = read.csv(input_path_R("Malignant_COI5_Angiogenesis_vs_Receiving_MI13_scatter.csv")),

  Inflammation_MI6   = read.csv(input_path_R("Malignant_COI5_Inflammation_vs_Receiving_MI6_scatter.csv")),
  Inflammation_MI10  = read.csv(input_path_R("Malignant_COI5_Inflammation_vs_Receiving_MI10_scatter.csv")),
  Inflammation_MI13  = read.csv(input_path_R("Malignant_COI5_Inflammation_vs_Receiving_MI13_scatter.csv"))
)

names(data_list)
```

# 5. Define the analysis plan

This table specifies which dataset to use, how the output file should be named, which candidate y-axis columns should be checked, and which color scheme should be applied.

```{r analysis-specs}
analysis_specs <- list(
  list(key = "CAF_MI10",          prefix = "CAF_vs_SendingMI10",            y_candidates = caf_y_candidates,   fill_low = caf_fill_low,   fill_high = caf_fill_high),
  list(key = "CAF_MI6",           prefix = "CAF_vs_SendingMI6",             y_candidates = caf_y_candidates,   fill_low = caf_fill_low,   fill_high = caf_fill_high),
  list(key = "CAF_MI13",          prefix = "CAF_vs_SendingMI13",            y_candidates = caf_y_candidates,   fill_low = caf_fill_low,   fill_high = caf_fill_high),

  list(key = "Metastasis_MI10",   prefix = "Metastasis_vs_ReceivingMI10",   y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Metastasis_MI6",    prefix = "Metastasis_vs_ReceivingMI6",    y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Metastasis_MI13",   prefix = "Metastasis_vs_ReceivingMI13",   y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),

  list(key = "Hypoxia_MI10",      prefix = "Hypoxia_vs_ReceivingMI10",      y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Hypoxia_MI6",       prefix = "Hypoxia_vs_ReceivingMI6",       y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Hypoxia_MI13",      prefix = "Hypoxia_vs_ReceivingMI13",      y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),

  list(key = "Angiogenesis_MI10", prefix = "Angiogenesis_vs_ReceivingMI10", y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Angiogenesis_MI6",  prefix = "Angiogenesis_vs_ReceivingMI6",  y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Angiogenesis_MI13", prefix = "Angiogenesis_vs_ReceivingMI13", y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),

  list(key = "Inflammation_MI6",  prefix = "Inflammation_vs_ReceivingMI6",  y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Inflammation_MI10", prefix = "Inflammation_vs_ReceivingMI10", y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high),
  list(key = "Inflammation_MI13", prefix = "Inflammation_vs_ReceivingMI13", y_candidates = state_y_candidates, fill_low = state_fill_low, fill_high = state_fill_high)
)
```

# 6. Run the analysis and save boxplots

For each comparison, the notebook will detect the relevant columns, compute summary statistics, generate the boxplot, save it as a PDF, and store the result for later review.

```{r run-analysis}
plot_results <- list()
all_boxplots <- list()
summary_list <- list()

for (spec in analysis_specs) {
  message("Running: ", spec$prefix)

  df <- data_list[[spec$key]]

  xcol <- pick_first_existing(df, x_candidates)
  ycol <- pick_first_existing(df, spec$y_candidates)

  stats_df <- summarize_two_groups(
    df = df,
    xcol = xcol,
    ycol = ycol,
    thr = thr_use
  ) %>%
    mutate(
      comparison = spec$prefix,
      .before = 1
    )

  p_box <- make_box_plot(
    df = df,
    xcol = xcol,
    ycol = ycol,
    thr = thr_use,
    fill_low = spec$fill_low,
    fill_high = spec$fill_high,
    title_txt = NULL,
    xlab = NULL,
    ylab = NULL
  )

  save_plot_pdf(
    p = p_box,
    fn = paste0(spec$prefix, "_box_wilcox_thr", thr_tag, ".pdf"),
    width = 1.2,
    height = 2.6
  )

  plot_results[[spec$key]] <- list(
    data = df,
    xcol = xcol,
    ycol = ycol,
    stats = stats_df,
    box = p_box
  )

  all_boxplots[[spec$prefix]] <- p_box
  summary_list[[spec$prefix]] <- stats_df
}
```

# 7. Summary table

This table provides a compact overview of all comparisons.

```{r summary-table}
summary_df <- bind_rows(summary_list) %>%
  mutate(
    mean_high = signif(mean_high, 4),
    mean_low = signif(mean_low, 4),
    mean_diff = signif(mean_diff, 4),
    log2_mean_change = signif(log2_mean_change, 4),
    p_wilcox = signif(p_wilcox, 4)
  )

kable(summary_df, align = "lcccccccccc")
```

# 8. Show all boxplots

This section prints all boxplots for quick inspection inside the notebook.

```{r show-plots, fig.width=2.2, fig.height=3.2}
for (nm in names(all_boxplots)) {
  message("Showing plot: ", nm)
  print(all_boxplots[[nm]])
}
```

# 9. Notes for users

## What usually needs to be changed

Use `HGSOC_LOCAL_RUN`, `HGSOC_RUN_DIR`, and `HGSOC_R_OUTPUT` to relocate files. Changing thresholds, file choices, or `analysis_specs` changes the analysis and is outside routine reproduction.

## Expected input format

Each CSV file should contain:

- one MI-related column, such as `MI_Level`, `Sending_MI10`, or `Receiving_MI10`
- one phenotype score column, such as `CAF_Score` or `State_Score`

## Output files

The notebook saves one PDF file per comparison to the following directory:

```{r output-path, echo=FALSE}
cat(file_savepath_main)
```

The output file names follow this pattern:

- `CAF_vs_SendingMI10_box_wilcox_thr0p50.pdf`
- `Metastasis_vs_ReceivingMI10_box_wilcox_thr0p50.pdf`

# Session information

```{r session-info}
sessionInfo()
```


# 10. Final 3 × 4 MI-program summary figure

This section generates one combined PDF figure matching the updated `HGSOC_Malignantsubtype_analysis` workflow. The rows are MI-10, MI-6, and MI-13. The columns are the four selected programs: CAF sig., PD-1/PD-L1 checkpoint, ECM-receptor interaction, and Hypoxia.

```{r final-3x4-mi-program-figure, fig.width=8.2, fig.height=7.2}
# =========================================================
# Final 3 × 4 MI-program boxplot figure
# ---------------------------------------------------------
# Rows:
#   MI-10, MI-6, MI-13
# Columns:
#   CAF sig., PD-1/PD-L1 checkpoint, ECM-receptor interaction, Hypoxia
#
# Annotation above each boxplot is: p-value stars + (SMD).
# SMD is the standardized mean difference between MI-high and MI-low groups:
#   SMD = (mean high - mean low) / pooled SD.
# Box fill and edge colors are high = #3D91CE and low = #81B6DE;
# all median lines are overlaid in white.
#
# Input CSVs expected from latest HGSOC_Malignantsubtype_analysis:
#   CAF:
#     Malignant_COI5_CAF_score_vs_Sending_MI{10,6,13}_scatter.csv
#   KEGG direct cell-level results:
#     Malignant_COI5_PD1_PDL1_vs_ReceivingMI{10,6,13}_celllevel_scatter.csv
#     Malignant_COI5_ECM_receptor_vs_ReceivingMI{10,6,13}_celllevel_scatter.csv
#   Hypoxia:
#     Malignant_COI5_Hypoxia_vs_Receiving_MI{10,6,13}_scatter.csv
#
# Output:
#   MI_CAF_functionalstate_3x4_CAF_PD1_ECM_Hypoxia_boxplot_SMD.pdf
#   MI_CAF_functionalstate_3x4_CAF_PD1_ECM_Hypoxia_boxplot_SMD.png
#   MI_CAF_functionalstate_3x4_CAF_PD1_ECM_Hypoxia_summary_SMD.csv
# =========================================================

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tibble)
})

# -----------------------------
# Colors matching the example figure
# -----------------------------
plot_colors <- c(
  "Sending MI-high"   = "#3D91CE",
  "Sending MI-low"    = "#81B6DE",
  "Receiving MI-high" = "#E83531",
  "Receiving MI-low"  = "#D9ADAC"
)

plot_group_order <- c(
  "Sending MI-high",
  "Sending MI-low",
  "Receiving MI-high",
  "Receiving MI-low"
)

# -----------------------------
# File-specification helpers
# -----------------------------
first_existing_file <- function(paths) {
  for (p in paths) {
    if (file.exists(p)) return(p)
  }
  return(NA_character_)
}

pick_first_existing_column <- function(df, candidates, file_label = "") {
  for (nm in candidates) {
    if (nm %in% colnames(df)) return(nm)
  }
  stop(
    paste0(
      "Cannot find any candidate column in ", file_label, ": ",
      paste(candidates, collapse = ", "),
      "\nAvailable columns: ", paste(colnames(df), collapse = ", ")
    )
  )
}

p_to_star_final <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-4) return("****")
  if (p < 1e-3) return("***")
  if (p < 1e-2) return("**")
  if (p < 0.05) return("*")
  return("ns")
}

# -----------------------------
# 3 × 4 panel specification
# -----------------------------
mi_order <- c("MI-10", "MI-6", "MI-13")
program_order <- c(
  "CAF sig.",
  "PD-1/PD-L1\ncheckpoint",
  "ECM-receptor\ninteraction",
  "Hypoxia"
)

panel_specs <- list(
  # MI-10 row
  list(
    mi_row = "MI-10", program = "CAF sig.",
    files = c(input_path_R("Malignant_COI5_CAF_score_vs_Sending_MI10_scatter.csv")),
    x_candidates = c("MI_Level", "Sending_MI10", "Sending_MI_10", "Sending MI10"),
    y_candidates = c("CAF_Score", "CAF_score", "State_Score", "state_score"),
    high_label = "Sending MI-high", low_label = "Sending MI-low"
  ),
  list(
    mi_row = "MI-10", program = "PD-1/PD-L1\ncheckpoint",
    files = c(
      input_path_R("Malignant_COI5_PD1_PDL1_vs_ReceivingMI10_celllevel_scatter.csv"),
      input_path_R("Malignant_COI5_PD1_PDL1_vs_Receiving_MI10_scatter.csv")
    ),
    x_candidates = c("MI_Level", "Receiving_MI10", "Receiving_MI_10", "Receiving MI10"),
    y_candidates = c("State_Score", "state_score", "pathway_module_score", "score"),
    high_label = "Receiving MI-high", low_label = "Receiving MI-low"
  ),
  list(
    mi_row = "MI-10", program = "ECM-receptor\ninteraction",
    files = c(
      input_path_R("Malignant_COI5_ECM_receptor_vs_ReceivingMI10_celllevel_scatter.csv"),
      input_path_R("Malignant_COI5_ECM_receptor_vs_Receiving_MI10_scatter.csv")
    ),
    x_candidates = c("MI_Level", "Receiving_MI10", "Receiving_MI_10", "Receiving MI10"),
    y_candidates = c("State_Score", "state_score", "pathway_module_score", "score"),
    high_label = "Receiving MI-high", low_label = "Receiving MI-low"
  ),
  list(
    mi_row = "MI-10", program = "Hypoxia",
    files = c(
      input_path_R("Malignant_COI5_Hypoxia_vs_Receiving_MI10_scatter.csv"),
      input_path_R("Malignant_COI5_Hypoxia_vs_ReceivingMI10_celllevel_scatter.csv")
    ),
    x_candidates = c("MI_Level", "Receiving_MI10", "Receiving_MI_10", "Receiving MI10"),
    y_candidates = c("State_Score", "state_score", "Hypoxia_score", "score"),
    high_label = "Receiving MI-high", low_label = "Receiving MI-low"
  ),

  # MI-6 row
  list(
    mi_row = "MI-6", program = "CAF sig.",
    files = c(input_path_R("Malignant_COI5_CAF_score_vs_Sending_MI6_scatter.csv")),
    x_candidates = c("MI_Level", "Sending_MI6", "Sending_MI_6", "Sending MI6"),
    y_candidates = c("CAF_Score", "CAF_score", "State_Score", "state_score"),
    high_label = "Sending MI-high", low_label = "Sending MI-low"
  ),
  list(
    mi_row = "MI-6", program = "PD-1/PD-L1\ncheckpoint",
    files = c(
      input_path_R("Malignant_COI5_PD1_PDL1_vs_ReceivingMI6_celllevel_scatter.csv"),
      input_path_R("Malignant_COI5_PD1_PDL1_vs_Receiving_MI6_scatter.csv")
    ),
    x_candidates = c("MI_Level", "Receiving_MI6", "Receiving_MI_6", "Receiving MI6"),
    y_candidates = c("State_Score", "state_score", "pathway_module_score", "score"),
    high_label = "Receiving MI-high", low_label = "Receiving MI-low"
  ),
  list(
    mi_row = "MI-6", program = "ECM-receptor\ninteraction",
    files = c(
      input_path_R("Malignant_COI5_ECM_receptor_vs_ReceivingMI6_celllevel_scatter.csv"),
      input_path_R("Malignant_COI5_ECM_receptor_vs_Receiving_MI6_scatter.csv")
    ),
    x_candidates = c("MI_Level", "Receiving_MI6", "Receiving_MI_6", "Receiving MI6"),
    y_candidates = c("State_Score", "state_score", "pathway_module_score", "score"),
    high_label = "Receiving MI-high", low_label = "Receiving MI-low"
  ),
  list(
    mi_row = "MI-6", program = "Hypoxia",
    files = c(
      input_path_R("Malignant_COI5_Hypoxia_vs_Receiving_MI6_scatter.csv"),
      input_path_R("Malignant_COI5_Hypoxia_vs_ReceivingMI6_celllevel_scatter.csv")
    ),
    x_candidates = c("MI_Level", "Receiving_MI6", "Receiving_MI_6", "Receiving MI6"),
    y_candidates = c("State_Score", "state_score", "Hypoxia_score", "score"),
    high_label = "Receiving MI-high", low_label = "Receiving MI-low"
  ),

  # MI-13 row
  list(
    mi_row = "MI-13", program = "CAF sig.",
    files = c(input_path_R("Malignant_COI5_CAF_score_vs_Sending_MI13_scatter.csv")),
    x_candidates = c("MI_Level", "Sending_MI13", "Sending_MI_13", "Sending MI13"),
    y_candidates = c("CAF_Score", "CAF_score", "State_Score", "state_score"),
    high_label = "Sending MI-high", low_label = "Sending MI-low"
  ),
  list(
    mi_row = "MI-13", program = "PD-1/PD-L1\ncheckpoint",
    files = c(
      input_path_R("Malignant_COI5_PD1_PDL1_vs_ReceivingMI13_celllevel_scatter.csv"),
      input_path_R("Malignant_COI5_PD1_PDL1_vs_Receiving_MI13_scatter.csv")
    ),
    x_candidates = c("MI_Level", "Receiving_MI13", "Receiving_MI_13", "Receiving MI13"),
    y_candidates = c("State_Score", "state_score", "pathway_module_score", "score"),
    high_label = "Receiving MI-high", low_label = "Receiving MI-low"
  ),
  list(
    mi_row = "MI-13", program = "ECM-receptor\ninteraction",
    files = c(
      input_path_R("Malignant_COI5_ECM_receptor_vs_ReceivingMI13_celllevel_scatter.csv"),
      input_path_R("Malignant_COI5_ECM_receptor_vs_Receiving_MI13_scatter.csv")
    ),
    x_candidates = c("MI_Level", "Receiving_MI13", "Receiving_MI_13", "Receiving MI13"),
    y_candidates = c("State_Score", "state_score", "pathway_module_score", "score"),
    high_label = "Receiving MI-high", low_label = "Receiving MI-low"
  ),
  list(
    mi_row = "MI-13", program = "Hypoxia",
    files = c(
      input_path_R("Malignant_COI5_Hypoxia_vs_Receiving_MI13_scatter.csv"),
      input_path_R("Malignant_COI5_Hypoxia_vs_ReceivingMI13_celllevel_scatter.csv")
    ),
    x_candidates = c("MI_Level", "Receiving_MI13", "Receiving_MI_13", "Receiving MI13"),
    y_candidates = c("State_Score", "state_score", "Hypoxia_score", "score"),
    high_label = "Receiving MI-high", low_label = "Receiving MI-low"
  )
)

# -----------------------------
# Load and harmonize all 12 panels
# -----------------------------
panel_data_list <- list()
panel_stats_list <- list()

for (i in seq_along(panel_specs)) {
  spec <- panel_specs[[i]]

  fn <- first_existing_file(spec$files)
  if (is.na(fn)) {
    stop(
      paste0(
        "Cannot find input file for ", spec$mi_row, " / ", spec$program,
        "\nTried:\n", paste(spec$files, collapse = "\n")
      )
    )
  }

  df_raw <- read.csv(fn, check.names = FALSE)
  xcol <- pick_first_existing_column(df_raw, spec$x_candidates, file_label = fn)
  ycol <- pick_first_existing_column(df_raw, spec$y_candidates, file_label = fn)

  df_panel <- df_raw %>%
    dplyr::mutate(
      MI_Level_for_plot = suppressWarnings(as.numeric(.data[[xcol]])),
      Score_for_plot = suppressWarnings(as.numeric(.data[[ycol]]))
    ) %>%
    dplyr::filter(is.finite(MI_Level_for_plot), is.finite(Score_for_plot)) %>%
    dplyr::mutate(
      MI_row = spec$mi_row,
      Program = spec$program,
      MI_group = ifelse(MI_Level_for_plot >= thr_use, spec$high_label, spec$low_label),
      xpos = ifelse(MI_Level_for_plot >= thr_use, 1, 2),
      input_file = fn,
      x_column = xcol,
      y_column = ycol
    )

  df_panel$MI_group <- factor(df_panel$MI_group, levels = plot_group_order)
  df_panel$MI_row <- factor(df_panel$MI_row, levels = mi_order)
  df_panel$Program <- factor(df_panel$Program, levels = program_order)

  high_vals <- df_panel$Score_for_plot[df_panel$xpos == 1]
  low_vals <- df_panel$Score_for_plot[df_panel$xpos == 2]

  p_wilcox <- NA_real_
  if (length(high_vals) > 0 && length(low_vals) > 0) {
    p_wilcox <- suppressWarnings(
      wilcox.test(high_vals, low_vals, alternative = "two.sided")$p.value
    )
  }

  mean_high <- mean(high_vals, na.rm = TRUE)
  mean_low <- mean(low_vals, na.rm = TRUE)
  mean_diff <- mean_high - mean_low

  sd_high <- stats::sd(high_vals, na.rm = TRUE)
  sd_low <- stats::sd(low_vals, na.rm = TRUE)
  n_high <- length(high_vals)
  n_low <- length(low_vals)

  pooled_sd <- NA_real_
  if (n_high > 1 && n_low > 1 && is.finite(sd_high) && is.finite(sd_low)) {
    pooled_sd <- sqrt(((n_high - 1) * sd_high^2 + (n_low - 1) * sd_low^2) / (n_high + n_low - 2))
  }

  smd <- ifelse(
    is.finite(mean_high) & is.finite(mean_low) & is.finite(pooled_sd) & pooled_sd > 0,
    (mean_high - mean_low) / pooled_sd,
    NA_real_
  )
  smd_label <- ifelse(is.finite(smd), sprintf("%.2f", smd), "NA")

  y_min <- min(df_panel$Score_for_plot, na.rm = TRUE)
  y_max <- max(df_panel$Score_for_plot, na.rm = TRUE)
  y_span <- max(y_max - y_min, 1e-6)

  panel_stats <- tibble::tibble(
    MI_row = spec$mi_row,
    Program = spec$program,
    n_high = n_high,
    n_low = n_low,
    mean_high = mean_high,
    mean_low = mean_low,
    mean_diff = mean_diff,
    sd_high = sd_high,
    sd_low = sd_low,
    pooled_sd = pooled_sd,
    SMD = smd,
    p_wilcox = p_wilcox,
    significance = p_to_star_final(p_wilcox),
    label = paste0(p_to_star_final(p_wilcox), " (SMD = ", smd_label, ")"),
    x = 1.5,
    x_start = 1,
    x_end = 2,
    y_start = y_max + 0.08 * y_span,
    y_text = y_max + 0.105 * y_span,
    y_tick = 0.04 * y_span,
    input_file = fn,
    x_column = xcol,
    y_column = ycol
  )

  panel_stats$MI_row <- factor(panel_stats$MI_row, levels = mi_order)
  panel_stats$Program <- factor(panel_stats$Program, levels = program_order)

  panel_data_list[[i]] <- df_panel
  panel_stats_list[[i]] <- panel_stats
}

combined_panel_df <- dplyr::bind_rows(panel_data_list)
combined_stats_df <- dplyr::bind_rows(panel_stats_list)

combined_panel_df$MI_row <- factor(combined_panel_df$MI_row, levels = mi_order)
combined_panel_df$Program <- factor(combined_panel_df$Program, levels = program_order)
combined_panel_df$MI_group <- factor(combined_panel_df$MI_group, levels = plot_group_order)

combined_stats_df$MI_row <- factor(combined_stats_df$MI_row, levels = mi_order)
combined_stats_df$Program <- factor(combined_stats_df$Program, levels = program_order)

# IMPORTANT:
# Use one facet label per MI × program panel and facet_wrap(scales = "free_y")
# below. ggplot2::facet_grid(scales = "free_y") still shares y scales within
# each row, so panels in the same MI row share y-axis ticks.
facet_panel_order <- unlist(lapply(mi_order, function(mi) paste(mi, program_order, sep = "\n")))

combined_panel_df <- combined_panel_df %>%
  dplyr::mutate(
    Facet_panel = factor(
      paste(as.character(MI_row), as.character(Program), sep = "\n"),
      levels = facet_panel_order
    )
  )

combined_stats_df <- combined_stats_df %>%
  dplyr::mutate(
    Facet_panel = factor(
      paste(as.character(MI_row), as.character(Program), sep = "\n"),
      levels = facet_panel_order
    )
  )

# Median segments are drawn manually in white so that every boxplot has
# a white median line while preserving high/low colored box edges.
combined_median_df <- combined_panel_df %>%
  dplyr::group_by(Facet_panel, MI_row, Program, xpos) %>%
  dplyr::summarise(
    median_score = median(Score_for_plot, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    x_start = xpos - 0.31,
    x_end = xpos + 0.31
  )

combined_median_df$MI_row <- factor(combined_median_df$MI_row, levels = mi_order)
combined_median_df$Program <- factor(combined_median_df$Program, levels = program_order)
combined_median_df$Facet_panel <- factor(combined_median_df$Facet_panel, levels = facet_panel_order)

summary_out <- file.path(
  file_savepath_main,
  paste0("MI_CAF_functionalstate_3x4_CAF_PD1_ECM_Hypoxia_summary_SMD_thr", thr_tag, ".csv")
)
write.csv(combined_stats_df, summary_out, row.names = FALSE)
message("Saved 3x4 summary table: ", summary_out)

# -----------------------------
# Generate final 3 × 4 figure
# -----------------------------
p_3x4 <- ggplot(combined_panel_df, aes(x = xpos, y = Score_for_plot, group = interaction(MI_group, xpos))) +
  geom_boxplot(
    aes(fill = MI_group, color = MI_group),
    width = 0.62,
    outlier.shape = NA,
    linewidth = 0.45
  ) +
  geom_segment(
    data = combined_median_df,
    aes(x = x_start, xend = x_end, y = median_score, yend = median_score),
    inherit.aes = FALSE,
    linewidth = 0.7,
    color = "white",
    lineend = "butt"
  ) +
  geom_segment(
    data = combined_stats_df,
    aes(x = x_start, xend = x_end, y = y_start, yend = y_start),
    inherit.aes = FALSE,
    linewidth = 0.45,
    color = "black"
  ) +
  geom_segment(
    data = combined_stats_df,
    aes(x = x_start, xend = x_start, y = y_start, yend = y_start - y_tick),
    inherit.aes = FALSE,
    linewidth = 0.45,
    color = "black"
  ) +
  geom_segment(
    data = combined_stats_df,
    aes(x = x_end, xend = x_end, y = y_start, yend = y_start - y_tick),
    inherit.aes = FALSE,
    linewidth = 0.45,
    color = "black"
  ) +
  geom_text(
    data = combined_stats_df,
    aes(x = x, y = y_text, label = label),
    inherit.aes = FALSE,
    size = 4.2,
    family = "Arial",
    fontface = "plain",
    color = "black"
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("", ""), limits = c(0.55, 2.45)) +
  scale_fill_manual(values = plot_colors, breaks = plot_group_order, drop = FALSE, name = NULL) +
  scale_color_manual(values = plot_colors, breaks = plot_group_order, drop = FALSE, name = NULL) +
  labs(x = NULL, y = "Mean expression") +
  coord_cartesian(clip = "off") +
  theme_classic(base_size = 13, base_family = "Arial") +
  theme(
    strip.placement = "outside",
    strip.background = element_rect(fill = "#ECECEE", color = NA),
    strip.text.x = element_text(size = 12.5, color = "black", margin = margin(t = 3, b = 3)),
    strip.text.y.left = element_text(size = 12.5, color = "black", angle = 90, margin = margin(r = 4, l = 4)),
    axis.title.y = element_text(size = 12, color = "black"),
    axis.text.y = element_text(size = 10, color = "black"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.55),
    axis.ticks.y = element_line(color = "black", linewidth = 0.55),
    panel.spacing.x = grid::unit(1.0, "lines"),
    panel.spacing.y = grid::unit(1.15, "lines"),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(size = 10.5, color = "black"),
    legend.key.size = grid::unit(0.45, "cm"),
    legend.spacing.x = grid::unit(0.35, "cm"),
    plot.margin = margin(t = 6, r = 8, b = 4, l = 4),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  guides(
    fill = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(linewidth = 0.7)),
    color = "none"
  ) +
  # facet_wrap gives each MI × program panel its own y scale.
  # Do not replace this with facet_grid(scales = "free_y"), because
  # facet_grid still shares y-axis limits/ticks within the same row.
  facet_wrap(~ Facet_panel, ncol = 4, scales = "free_y")

pdf_out <- file.path(
  file_savepath_main,
  paste0("MI_CAF_functionalstate_3x4_CAF_PD1_ECM_Hypoxia_boxplot_SMD_thr", thr_tag, ".pdf")
)
png_out <- file.path(
  file_savepath_main,
  paste0("MI_CAF_functionalstate_3x4_CAF_PD1_ECM_Hypoxia_boxplot_SMD_thr", thr_tag, ".png")
)

if (capabilities("cairo")) {
  ggsave(pdf_out, plot = p_3x4, width = 8.2, height = 7.2, device = cairo_pdf, bg = "white")
} else {
  ggsave(pdf_out, plot = p_3x4, width = 8.2, height = 7.2, bg = "white")
}

ggsave(png_out, plot = p_3x4, width = 8.2, height = 7.2, dpi = 300, bg = "white")

message("Saved 3x4 PDF: ", pdf_out)
message("Saved 3x4 PNG: ", png_out)

print(p_3x4)
```
